---
title: "Creating words from iterated vocal imitation"
bibliography: telephone.bib
output:
  pdf_document:
    citation_package: natbib
    template: templates/cogsci.tex
abstract: |
  We report the results of a large-scale (_N_ = 1402) experiment to investigate whether spoken words can emerge from the process of repeated imitation. Participants played a version of the children’s game “Telephone”. The first generation were asked to to imitate recognizable environmental sounds (e.g., glass breaking, water splashing); subsequent generations imitated the imitators, for a total of 8 generations. We then examined whether the vocal imitations became more stable and word-like, retained a resemblance to the original sound, and were suitable as learned category labels. The results showed (1) the imitations became progressively more word-like, (2) even after 8 generations, could be matched above chance to the environmental sound that motivated them, and (3) imitations from later generations were more effective as learned category labels. These results show how repeated imitation can create progressively word-like forms while retaining a semblance of iconicity.
---

Over the ages, people have pondered the origins of languages and especially the
words that compose them (e.g., Plato, Bible). Some theories of language
evolution hypothesize that vocal imitation played an important role in
generating the first words of spoken languages [@Brown:1955wy; @Donald:2016kd;
@Imai:2014dea; @Perlman:2015ip]. In this study, we investigate the formation of
onomatopoeic words--imitative words that resemble the sounds to which they
refer. We ask whether onomatopoeic words can be formed gradually and without
instruction simply from repeating the same imitation over generations of
speakers.

Onomatopoeic words appear to be a universal lexical category found across the
world’s languages [@Dingemanse:2012fc]. Languages all have conventional words
for animal vocalizations and other environmental sounds. Such words appear to be
especially prominent for very young language learners [@Laing:2016bo;
@Tardif:2008gg]. In some cases, words that begin as imitations of sounds become
fully lexicalized and integrated into the broader linguistic system, when they
behave like more “ordinary” words that can undergo typical morphological
processes. [@Dingemanse:2016dg] show this on a variable instance-by-instance
basis rather than a monotonic historical process towards integration.
[@Rhodes:1994au] observed that onomatopoeic words exist along a continuum from
“wild” to “tame” in which wild words have a more imitative phonology whereas
tame words take on more standard phonology of other English words. In English,
words like "crack" or the recently adapted "ping".

In addition to iconic words for sounds, many languages have a semantically
broader system of ideophones. [@Dingemanse:2016vd] found that naïve listeners
were better than chance at guessing the meanings of ideophones (sampled from
five different languages) from five different semantic categories: color/visual,
motion, shape, sound, and texture, but sound words were guessed more accurately
than the rest. Experiments suggest that iconic vocalizations for non-sound
concepts can also become more word-like over repeated use [@Perlman:2015ip].

But the process by which onomatopoeic words may be born of nonverbal imitations
has yet to be observed in a controlled lab environment. Do repeated nonverbal
imitations become more wordlike even without any instruction to do so? Or
alternatively does the limited fidelity of human vocal imitation restrict such
stability from forming in the absence of a reason to communicate. To test this,
we recruited participants to engage in a large scale online version of the
children's game of "Telephone" in which an acoustic message is passed from one
person to the next. After obtaining these imitations, we then investigated how
the imitations changed over generations. We investigated the acoustic properties
of the imitations as well as the orthographic properties of the imitations once
transcribed into English words. Finally, we test how quickly these invented
words are learned as category labels in a category learning experiment.

```{r config, include=FALSE}
library(knitr)
library(tidyverse)
library(lme4)

opts_chunk$set(
  echo=FALSE,
  message=FALSE,
  warning=FALSE,
  results="hide",
  fig.path="figs/",
  out.width="8cm",  # 8.7 is max
  # out.height="5cm",
  dpi=144,
  cache=TRUE,
  cache.path=".cache/"
)

library(wordsintransition)

read_chunk("creating-words.R")
```

```{r setup, include=FALSE}
```

# Methods

The design of the transmission chain experiment is shown in Fig. 1. We
started with 16 seed sounds, four from each of four categories of environmental
sounds (seed selection procedures are described below). Four participants
imitated each environmental sound, starting unique transmission chains that were
allowed to continue for a maximum of 8 generations. We removed all low quality
recordings and recordings where the speaker violated the rules of the
experiment, e.g., by recording something in English, resulting in a final sample
of `r n_final_imitations` imitations along `r n_branches` contiguous
transmission chains.

```{r fig1, fig.width=4, out.width="4.0cm", fig.align="center", fig.cap="The design of the transmission chain experiment. 16 seed sounds were used, 4 in each category of nonverbal environmental sounds. Each seed sound was imitated by 4 different participants, resulting in 4 branches off of each seed sound. Subsequent participants imitated the imitations and so on for a maximum of 8 generations."}
draw_graphviz("definitions", "wordsintransition")
```

## Selecting seed sounds

We initially selected a set of 36 sounds in 6 different categories of
environmental sounds. Inanimate categories of sounds were selected because
they were less likely to have lexicalized onomatopoeic forms already in
English, and they were assumed to be less familiar and harder to imitate. We
reduced this initial set to the final sample of 16 seeds using an odd-one-out
norming procedure (_N_ = 105 participants) designed to remove any sounds
that were outliers within each category. The results of this norming
procedure and selection criteria are provided in the Supplemental Materials.

## Collecting vocal imitations

To collect the vocal imitations, participants were paid to participate in an 
online version of the children's game of "Telephone". The instructions stated
that participants would be imitating nonverbal sounds and that they should try
to reproduce the message as accurately as possible. Participants in this online
study listened to 4 messages and recorded 4 imitations from their homes or
offices using their personal recording equipment (usually a built-in
microphone). Participants were assigned one message from each of the four
categories of sounds at random so participants were unlikely to hear the same
person more than once.

Imitations were monitored as they were received by an experimenter. The purpose
of this monitoring was to catch any gross errors in recording before they were
passed on to the next generation of imitators. For example, most sounds required
trimming the recording more closely to the vocalization--removing mouse clicks
and other ambient sounds. Sounds were also prevented from being reproduced if
the imitator violated the rules of the experiment, e.g., by saying something in
English.

## Matching imitations to seeds

To measure the extent to which the imitations could be matched back to the
original seeds, a new set of participants listened to the imitations and guessed
which seed they came from. Each participant was assigned 4 different seed sounds
to use as choices in a 4AFC task. On each trial, they listened to an imitation,
and decided which of the 4 seed sounds they thought the imitation most closely
resembled. They did not receive any feedback on their performance. Each
participant also received a "catch" imitation, instructing them to select the
third option on that question. Responses from participants who failed any catch question were
excluded.

```{r fig2, fig.width=7, out.width="7.0cm", fig.align="center", fig.cap='Three types of matching questions depicted in relation to the original set of 16 seed sounds. For each question, participants listened to a sample imitation (orange circle) and had to guess which of 4 sound choices (green circles) they thought the person was trying to imitate. (A) True seed questions contained the actual seed that generated the imitation in the choices, and the distrator seeds were sampled from different categories. (B) Category match questions also used distractor sounds from different categories but the correct seed was not the actual seed, but a different sound within the same category. (C) Specific match questions pitted the actual seed against the other seeds within the same category.'}
q_true_seed <- read_graphviz("true-seed", "wordsintransition")
q_category_match <- read_graphviz("category-match", "wordsintransition")
q_specific_match <- read_graphviz("specific-match", "wordsintransition")

grid.arrange(
  q_true_seed,
  q_category_match,
  q_specific_match,
  ncol = 1
)
```

## Collecting transcriptions of imitations

Participants were paid to transcribe imitations into written words. We selected
the first imitation in each transmission chain as well as the three most distal
messages for transcription. Only the transcriptions from participants who passed
a catch question were included in the analysis.

## Matching transcriptions to seeds

Participants completed a modified version of the matching game. Instead of
listening to imitations, participants now read a transcription of an imitation.
They were instructed that the person who invented that word was talking about
one of four seed sounds presented to each participant, and they had to guess
which one. We restricted the size of the experiment by using only the 4 most
frequent transcriptions for each imitation.

## Using transcriptions as category labels

We selected a set of transcriptions to use as novel category labels in a simple
category learning experiment. To determine which transcriptions to test in the
experiment, we first selected only those transcriptions which had above chance
matching performance when matching back to the original seeds. Then we excluded
transcriptions that had less than 2 unique characters, and transcriptions that
were over 10 characters long. From this set of eligible transcriptions, we
sampled from both first and last generation imitations to reach a final set that
was controlled for overall matching accuracy.

# Results

## Changes in imitation fidelity

```{r collecting-imitations, include=FALSE}
```

Imitations from later generations were rated as being more similar to one another than imitations from earlier generations, `r lmer_mod_results(similarity_judgments_mod, "edge_generation_n")` (**Fig. 2**). Given the nonverbal nature of the imitations, large differences between speakers (e.g., speakers of different genders), noise in the recordings, and variability in recording equipment from conducting the experiment online, previously published techniques for calculating acoustic distance were inadequate [cf. @Lemaitre:2016kz]. Instead, we obtained subjective measures of acoustic similarity using a controlled, randomized norming procedure (see Methods). We found that imitations from later generations where rated as being more similar to one another than imitations from earlier generations, suggesting that the acoustic signals were becoming more stable through repetition.

```{r fig3, fig.cap="Increase in acoustic similarity over generations."}
gg_similarity_judgments
```

## Matching imitations to original sounds

```{r matching-imitations, include=FALSE}
```

Matching accuracy for all question types was above chance for first generation imitations, `r glmer_mod_results(imitation_matches_overall_mod, "(Intercept)", odds = TRUE)`, and decreased over generations, `r glmer_mod_results(imitation_matches_overall_mod, "generation_1")`. We tested whether this decay in match accuracy was constant across the three question types or if some question types were more or less resistant to generational decay than others. In particular we hypothesized that if the imitations were becoming more word-like and therefore categorical as they were repeated, then performance on questions where category information enabled a correct response will be more resilient to generational decay.

The results are shown in **Fig. 4**. The first evidence in support of our hypothesis comes in comparing performance on questions requiring a category match to performance on questions where guessing correctly required distinguishing the true seed from other sounds within the same category. Performance decreased over generations more rapidly for these specific match questions than for category match questions, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_between")`, suggesting that category information was more resistent to loss through transmission. A simpler explanation for this result is that the specific match questions are simply harder than the category match questions. However, performance also decreased more rapidly for the easiest type of question where the correct answer was the actual seed generating the imitation. The advantage for having the true seed among the options decreased over generations, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_within")`. These results indicate that later generation imitations were more likely to be recognized as identifiers of a particular category than they were of particular exemplars within each category.

```{r fig4, fig.cap='Matching accuracy for guessing the seed from imitations taken from different generations of the transmission chain experiment. Lines denote different question types concerning the relationship between the imitation and the options in the question (see Fig. 3). The advantage of having a True seed among the options (True seed versus Category match) decreased over generations, and the advantage of having distractors from different categories (Category match versus Specific match) increased over generations. These results suggest that category information was more resilient to generational decay than specific information about the particular exemplar being imitated.'}
gg_match_to_seed
```

## Changes in orthographic agreement

```{r transcriptions, include=FALSE}
```

We next investigated how the orthographic forms of the imitations changed over generations. A total of `r n_imitations_transcribed` imitations were transcribed, with approximately `r n_transcriptions_per_imitation` transcriptions per imitation. Analyzing changes in orthographic agreement over generations paralleled what was observed in the analysis of acoustic similarity: Transcriptions from later generation imitations were more similar to one another in terms of orthographic distance than transcriptions from earlier generations, `r lmer_mod_results(orthographic_distance_mod, "message_c")` (**Fig. 5**). This result supports the conclusion that the imitations were becoming more stable in both acoustic and orthographic forms.

```{r fig5, fig.cap="Average orthographic distance among transcriptions of imitations taken from first and last generations."}
gg_distance
```

## Matching imitations back to original sounds

To investigate whether these invented words retained any resemblance to the original seed sounds, we selected the top 4 most frequent transcriptions for each imitation that was transcribed and presented them in a modified version of the "Guess the seed" matching game (_N_ = `r n_transcription_match_subjs`). On each trial, participants were presented with a random transcription generated from one of the imitations, and tasked with guessing the meaning of the invented word from among 4 possible sounds. Participants were able to guess the correct meaning of the transcribed word above chance even after between 5 and 8 generations of repetition, `r glmer_mod_results(transcription_matches_last_gen_mod, "(Intercept)", odds = TRUE)` (**Fig. 6**). This was true both for true seed questions, `r glmer_mod_results(transcription_matches_last_gen_mod_category, "(Intercept)", odds = TRUE)`, and for category match questions, `r glmer_mod_results(transcription_matches_last_gen_mod_exact, "(Intercept)", odds = TRUE)`. The effect of generation did not vary across these question types, `r glmer_mod_results(acc_mod, "question_c:message_c")`.

```{r fig6, fig.cap="Matching accuracy for transcriptions of imitations taken from first and last generations. True seed questions contained transcriptions of the actual seed generating the transcribed word. Category match questions contained transcriptions of imitations of other seeds from the same category."}
gg_match_transcriptions
```

The result that matching accuracy was above chance even for category match questions is especially informative. This finding indicates that participants were able to match a transcription back to a seed sound that was from the same category, but not the exact chain, as the original sound leading to the transcription. If transcriptions only captured idiosyncratic elements of the imitations, then the only way participants would be able to guess the "meaning" of the transcription at above chance levels would be through its surface resemblance to the specific sound being transcribed, in which case performance in category match questions would be indistinguishable from chance. Instead we found that the transcription--the result of an imitation passed through generations of different speakers--still retained some resemblance to the category of sounds being imitated, and thus participants were able to use this category information to match the transcribed word to the category of sounds at above chance levels. This result supports our hypothesis that through repetition, imitations are becoming more wordlike.

## Using transcriptions as novel category labels

```{r category-learning}
```

Last, we examined whether there was a learning advantage to the more wordlike imitations emerging through iterated repetition compared to direct imitations of the source of the sound. The advantage of word cues as opposed to environmental cues to a particular category is that words activate more categorical representations useful in generalizing across idiosyncratic differences between category members [@Edmiston:2015he]. Therefore, we predicted that transcriptions of the more wordlike forms emerging through repeated imitation should be easier to learn as category labels requiring generalization to new category members than transcriptions from direct imitations.

To test this, we recruited participants (_N_ = `r n_lsn_subjs`) to learn a novel category label for each of the 4 categories of environmental sounds used in the transmission chain experiment. When participants had to generalize the meaning of the novel label to new category members (new sounds), they were faster when the label came from transcriptions of later generation imitations than from transcriptions of first generation imitations, `r lmer_mod_results(lsn_after_first_block_mod, "message_c")`. This suggests that in addition to becoming more stable both in terms of acoustic and orthographic properties, imitations that have been more repeated are also easier to learn as category labels.

```{r fig7, fig.cap="Response times across blocks in the category learning experiment. Each new block introduced four new sounds into the categories participants were learning the names of. Participants who learned labels that were actually transcriptions from the end of the transmission chain branches were faster to generalize these labels to new category members than participants who learned labels transcribed directly from imitations of the source sounds."}
rt_plot
```

# Discussion

We show that repeated imitation of an originally imitative vocalization works to integrate it into the ambient language. Consequently, it becomes more abstract in its resemblance to its initial source sound, but maintains some resemblance to the more abstract category of sounds. The vocalizations become more learnable as linguistic labels for categories.

Repeated drawings from memory (without delay) result in drawings that become more conventional, moving towards cultural attractors such as letters [@Tamariz:2014gr]. Also Bartlett.

completing the transition from nonverbal imitation to a fully lexicalized word form and demonstrating the impact of this transition on communication

More broadly, imitative words are hypothesized to play an important role in the evolution of language [@Brown:1955wy; @Imai:2014dea; @Perlman:2015ip].

# References
