---
title: "Creating words from iterated vocal imitation"
bibliography: telephone.bib
output:
  pdf_document:
    citation_package: natbib
    template: templates/pnas.tex
    keep_tex: true
abstract: |
  We report an investigation into how spoken words might emerge from the repeated imitation of environmental sounds. Participants played a version of the children’s game “Telephone”. The first generation imitated recognizable environmental sounds (e.g., glass breaking, water splashing) and subsequent generations imitated the imitations of the prior generation for a total of 8 generations. We then examined whether the vocal imitations became more word-like, became more suitable as learned category labels, and retained a resemblance to the original sound. The results showed that the imitations became more stable and word-like, and more easily learned as category labels. At the same time, even after 8 generations, both spoken imitations and their written transcriptions could be matched above chance to the category of environmental sound that motivated them. These results show how repeated imitation can create progressively more word-like forms while retaining a semblance of iconicity with the original sound.
---

```{r config, include=FALSE}
library(knitr)

opts_chunk$set(
  echo=FALSE,
  message=FALSE,
  warning=FALSE,
  results="hide",
  fig.path="figs/",
  fig.width=7.7,
  fig.height=5.5,
  fig.align="center",
  out.width="7.7cm",  # max is 8.7cm
  dpi=144,
  cache=TRUE,
  cache.path=".cache/",
  autodep=TRUE
)

read_chunk("creating-words.R")
read_chunk("R/dendrogram.R")
```

```{r setup, include=FALSE, cache=FALSE}
```

People have long pondered the origins of languages, especially the words that compose them. For example, both Plato in his _Cratylus_ dialogue [@Plato:1999uk] and John Locke in his _Essay Concerning Human Understanding_ [@Locke:1948eu] examined the "naturalness" of words--whether they are somehow imitative of their meaning. Here, we report a large-scale experiment investigating whether new words can form--gradually and without instruction--simply by repeating imitations of environmental sounds.

The importance of imitation and depiction in the origin of signs is clearly observable in signed languages [@Klima:1979], but in considering the idea that vocal imitation may be key to understanding the origin of spoken words, many have argued that the human capacity for vocal imitation is far too limited [@Arbib:2012; @Armstrong:2008; @Corballis:2002; @Hocket:1978; @Tomasello:2008]. For example, [@Pinker:2005cv] argued that, “most humans lack the ability (found in some birds) to convincingly reproduce environmental sounds … Thus ‘capacity for vocal imitation’ in humans might be better described as a capacity to learn to produce speech” (p. 209). Consequently, it is still widely assumed that vocal imitation--or more broadly, the use of any sort of resemblance between form and meaning--cannot be important to understanding the origin of spoken words [@GoldinMeadow:2016; @Kendon:2014].

But although most words of contemporary spoken languages are not clearly imitative in origin, there has been a growing recognition of the preponderance of imitative words in spoken language [@Dingemanse:2015cu; @Perniss:2010fb] and the frequent use of vocal imitation and depiction in spoken discourse [@Clark:2016; @Lewis:2009], and some researchers have argued for the importance of imitation for understanding the origin of spoken words [e.g., @Brown:1955wy; @Donald:2016kd; @Imai:2014dea; @Perlman:2015ip; @Dingemanse:2014gj]. In addition, experiments showing that people can, in fact, be highly effective at using vocal imitations to refer to different kinds of sounds – in some cases, even more effective than with the use of words [@Lemaitre:2014kr]. The effectiveness of these imitations arises not because people are able to mimic environmental sounds with high-fidelity, but because they are able to represent the salient features of sounds in ways that are understandable to listeners [@Lemaitre:2016kz]. Similarly, the features of onomatopoeic words might highlight distinctive aspects of the sound it represents: for example, the initial voiced, plosive /b/ in “boom” represents an abrupt, loud onset, the back vowel /u/ a low pitch, and the nasalized /m/ a slow, muffled decay [@Rhodes:1994au]. Recent work has also shows that people are able to create novel imitative vocalizations for more abstract meanings (e.g. ‘slow’, ‘rough’, ‘good’, ‘many’), in ways that are understandable to naïve listeners [@Perlman:2015ip].

Thus, research shows that people can use vocal imitation as an effective means to communicate about the various sounds of their environment and even more abstract concepts. But how do vocal imitations become standardized words that are integrated into the vocabulary of a language? To investigate this question, we recruited participants to play an online version of the children's game of "Telephone". In the children’s game, a spoken message is whispered from one person to the next. In our version, the original message was a recording of an environmental sound. The first generation participant imitated the sound, the next generation imitated the previous imitation, and so on for up to 8 generations.

In a series of subsequent studies, we systematically answered the following questions. First, does iterated imitation drive the vocalizations to stabilize in form and become more word-like? Second, do the imitations become more suitable as labels for the category of sounds that motivated them? For example, does the imitation of a particular water-splashing sound become, over time, a better label for the more general category of water-splashing sounds? Third, do the imitations retain resemblance to the original environmental sounds that inspired them? If so, it should be possible for naïve participants to match the emergent imitative words back to the original sounds that motivated them.

```{r dendrogram, fig.width=11.4, fig.out="11.4cm", fig.align="center", fig.cap="The design of the transmission chain experiment. 16 seed sounds were selected, four in each category of environmental sound. Participants imitated each seed sound, and then the next generation of participants imitated the imitations and so on for 8 generations."}
```



# Results {-}

We begin with a summary of our main results. Measuring the acoustic similarity of repeated imitations revealed that imitations became more similar to one another through repetition. As the imitations were repeated, they gradually lost their resemblance to the source sound. In particular, they lost information that distinguished the source sound from within-category competitors more readily than higher-level category information. This result suggests that through repetition and stabilization the imitations became better abstract category labels by virtue of cueing all category members equally as opposed to highlighting individual category members. We found support for this conclusion in the transcriptions of the imitations. Later generations of imitations were transcribed with better agreement, suggesting that imitations were indeed stabilizing on invented words that were increasingly distinctive and broadly recognizable. Still, these invented words retained some resemblance to the category of environmental sound that motivated them (at least relative to the other categories tested in this experiment). Participants were able to accurately match the transcriptions of final generation imitations in each transmission chain back to the category of environmental sounds that motivated them. Unlike the direct matching of imitations, the extent to which transcriptions were matched to individual source sounds as opposed to categories of sounds did not increase over generations. However, when transcriptions of first and last generation imitations were learned as novel labels of environmental sound categories, last generation transcriptions were easier to learn than those from the first generation. These results describe a process by which an imitation of an environmental sound may transition to a more word-like form through unguided repetition, and suggest that such a transition to more word-like forms might make them more effective as category labels.

## Imitations became more stable and word-like

```{r collecting-imitations, include=FALSE}
```

We collected a total of `r n_all_imitations` imitations from `r n_imitators` participants in a study conducted online. `r n_removed` imitations were removed for bad audio quality or violating the rules of the experiment (e.g., saying something in English), leaving `r n_final_imitations` imitations along `r n_branches` contiguous transmission chains for analysis.

Trained research assistants coded these imitations for acoustic similarity using a blinded, pairwise comparison procedure (see Methods). Inter-rater reliability was high, `r report_icc_results(irr_results)`. Similarity ratings were fit with a hierarchical linear model predicting similarity from generation with random effects for rater and for category. Imitations from later generations were rated as sounding more similar to one another than imitations from earlier generations, `r lmer_mod_results(similarity_judgments_lmertest_mod, "edge_generation_n")` (Fig. 2). This result suggests that imitations may be stabilizing on particular acoustic forms through repetition.

```{r fig2, fig.height=5.5, fig.cap="Change in acoustic similarity over generations of repetition. Points show mean acoustic similarity ratings for imitations in each category of environmental sounds. The line shows the linear predictions of a hierarchical model with random effects for rater and category, with error bands designating +/- 1 standard error of the model predictions. The results show that acoustic similarity increases over generations, indicating that subsequent imitations become more similar to one another through repetition."}
gg_similarity_judgments
```

We also calculated automated analyses of imitation fidelity using Mel Frequency Cepstral Coefficients (MFCCs) as a measure of acoustic distance. However, for our stimuli the correlation between automated analyses of acoustic similarity and rater judgments was low, `r report_cor_test(similarity_cor_test)`, suggesting that the automated analyses do not capture the acoustic features driving the perception of acoustic similarity. This is possibly due to the non-verbal nature of the imitations as well as variation in recording quality between participants in the online study. We therefore report the results of these automated analyses in the Supporting Information.

## Transcription agreement increased over generations

```{r transcriptions, include=FALSE}
```

We next tested whether the imitations became more clearly distinguishable as particular words, as opposed to non-linguistic, i.e., non-English sounds. We had English-speaking participants transcribe the imitations into English orthography, and then we measured whether transcription agreement increased over generations. We selected the first and final three imitations in each transmission chain to be transcribed. As a control, we also obtained "transcriptions" of the seed sounds themselves. `r n_transcribers` participants generated a total of `r n_transcriptions` or approximately `r n_transcriptions_per_imitation` transcriptions per sound (imitation and seed sounds). Transcriptions containing actual English words and those from participants who failed a catch question were excluded from analysis (`n_transcriptions_dropped`).

```{r table1, results='asis'}
table_caption <- "Examples of invented words."
table_format <- opts_knit$get("out.format")

if(table_format == "markdown") {
  # If it's markdown, you're probably going to tex
  table_format <- "latex"
}

kable(transcription_examples, caption = table_caption,
      format = table_format, booktabs = TRUE)
```

To measure transcription agreement we took the average orthographic distance (longest contiguous matching subsequence) between the most frequent transcription and all other transcriptions of a given imitation. Hierarchical linear models were fit predicting orthographic distance from the type of imitation being transcribed (First generation imitations, Last 3 generation imitations) with random effects for transmission chains nested within categories.

Transcriptions of later generation imitations were more similar to one another in terms of orthographic distance than transcriptions from earlier generations, `r lmer_mod_results(orthographic_distance_lmertest_mod, "message_c")` (Fig. 5). This result supports our hypothesis that unguided repetition drives imitations to become more distinctive as particular English words. The same conclusion was reached from alternative measures of orthographic distance, including exact string matches and excluding those imitations for which all transcriptions were unique.

```{r fig5, fig.cap="Average orthographic distance among transcriptions of imitations taken from first and last generations. Each point shows the average orthographic distance between the most frequent transcription and all other transcriptions of a single imitation. Error bars are +/- 1 standard error of the linear mixed effects model predictions. Transcriptions of later generation imitations were more similar to one another than transcriptions of first generation imitations."}
gg_distance
```

## Repeated imitations were easier to learn as category labels

```{r category-learning, include=FALSE}
```

Our hypothesis was that repetition of imitations would result in increasingly word-like forms, but what are the consequences of this transition for the language user? To examine this question, we tested whether the words created through repetition were easier to learn as category labels.

When participants learned some of the transcriptions as novel category labels for categories of environmental sounds, they were faster when the label came from transcriptions of later generation imitations than from transcriptions of first generation imitations, `r lmer_mod_results(lsn_after_first_block_lmertest_mod, "message_c")` (Fig. 7A). In addition to becoming more stable both in terms of acoustic and orthographic properties, imitations that have been more repeated were also easier to learn as category labels.

The effect can be further localized within each block. Comparing RTs on the trials leading up to a block transition and the trials immediately after the block transition revealed a reliable interaction between block transition and the generation of the transcribed label, `r lmer_mod_results(transition_lmertest_mod, "block_transition_c:message_c")` (Fig. 7B). This result suggests that learning transcriptions from later generation imitations were easier to generalize to new category members.

```{r fig7, fig.height=11, out.height="11cm", fig.cap="(Top) RTs on correct trials by block, showing faster responses when learning category labels transcribed from last generation imitations. (Bottom) RTs on trials leading up to and immediately following the block transition where new category members are introduced."}
grid.arrange(rt_plot, gg_transition, ncol = 1)
```

## Imitations retained seed category information

```{r matching-imitations, include=FALSE}
```

Were the imitations stabilizing on arbitrary acoustic forms or were they maintaining some aspect of the original environmental sound? To test this, we measured the ability of participants naive to the design of the Telephone game to match each imitation back to its original source relative to other seed sounds from either the same category or from different categories (Fig. 3). All 365 imitations were tested in the three conditions depicted in Fig. 3. These conditions differed in the relationship between the imitation and the four seed sounds serving as the choices in the 4 alternative forced choice (4AFC) task. Responses were fit by hierarchical generalized linear models predicting match accuracy as different from chance (25%) based on the type of question being answered (True seed, Category match, Specific match) and the generation of the imitation.

```{r fig3, fig.width=8, out.width="8.0cm", fig.align="center", fig.cap='Types of 4AFC matching questions depicted in relation to the original set of 16 seed sounds. For each question, participants listened to an imitation (orange dashed circle) and had to guess which of 4 sound choices (green solid circles) they thought the person was trying to imitate. (Top) True seed questions contained the actual sound that generated the imitation in the choices, and the three distractor sounds were sampled from different categories. (Middle) Category match questions also used distractor sounds from different categories but the "correct" sound was not the actual seed, but a different sound within the same category. (Bottom) Specific match questions pitted the actual seed against the other seeds within the same category.'}
grid.arrange(
  q_true_seed,
  q_category_match,
  q_specific_match,
  ncol = 1
)
```

Matching accuracy for all question types started above chance for the first generation of imitations, `r glmer_mod_results(imitation_matches_overall_mod, "(Intercept)", odds = TRUE)`, and decreased steadily over generations, `r glmer_mod_results(imitation_matches_overall_mod, "generation_1")`. We tested whether this increase in matching difficulty was constant across the three types of questions or if some question types became more difficult at later generations than others. The results are shown in Fig. 4. Performance decreased over generations more rapidly for questions requiring a within-category distinction than for between-category questions, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_within")`, suggesting that between-category information was more resistant to loss through transmission. One explanation for this result is that the within-category match questions are simply more difficult because the sounds are more acoustically similar to one another than the between-category questions and therefore performance might be expected to drop off more rapidly with repeated imitations. However, performance also decreased for the easiest type of question where the correct answer was the actual seed generating the imitation (True seed questions; see Fig. 3); the advantage of having the true seed among between-category distractors decreased over generations, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_between")`.

```{r fig4, fig.height=5.5, fig.cap='Changes in matching accuracy over generations. Matching accuracy is the ability to guess the sound most likely to have generated the imitation relative to other seed sounds used in the experiment. Performance is separated by question type which describes the relationship between the imitation and the choices in the 4AFC task (see Fig. 3). Lines show predictions from a generalized linear mixed effects model along with +/- 1 standard error of the model predictions. The "category advantage" (category match v. specific match) increased over generations, while the "true seed advantage" (true seed v. category match) decreased. These results suggest that imitions lose within-category information more rapidly than between-category information.'}
gg_match_to_seed
```

These results indicate that as imitations are repeated they lose within-category information more rapidly than between-category information. Later generation imitations were just as likely to be recognized as identifiers of an entire category of environmental sounds as they were of particular sounds within the category.

## Transcriptions retained seed category information

```{r matching-transcriptions, include=FALSE}
```

We previously demonstrated that people were able to accurately guess the source of an imitation after 8 repetitions, but what about the source of a transcription of an imitation? Do these invented words still resemble the category of sounds that was originally imitated? We tested the top 4 most frequent transcriptions for each imitation in a modified version of the "Guess the Seed" game (see Fig. 3). Participants were given a novel word and had to guess which sound they thought the person who invented the word was talking about. The distractors for all questions were between-category, i.e. Specific match questions were not tested with transcriptions.

Participants were able to guess the correct meaning of the transcribed word above chance even after 8 generations of repetition, `r glmer_mod_results(transcription_matches_last_gen_mod, "(Intercept)", odds = TRUE)` (Fig. 6). This was true both for "True seed" questions containing the actual seed generating the transcribed imitation, `r glmer_mod_results(transcription_matches_last_gen_mod_category, "(Intercept)", odds = TRUE)`, and for "Category match" questions where participants had to associate transcriptions with a particular category of environmental sounds, `r glmer_mod_results(transcription_matches_last_gen_mod_exact, "(Intercept)", odds = TRUE)`.

Interestingly, the effect of generation did not vary across these question types, `r glmer_mod_results(acc_mod, "question_c:message_c")`. This indicates that transcriptions of imitations may capture idiosyncratic elements of specific category members more than the imitations themselves. Possible reasons for this asymmetry between imitations and transcriptions are explored in the Discussion.

```{r fig6, fig.cap="Matching accuracy for transcriptions of imitations taken from first and last generations. Bars represent the predicted means taken from the generalized linear mixed effects model with +/- 1 standard error of the model predictions. True seed questions contained transcriptions of the actual seed generating the transcribed word. Category match questions contained transcriptions of imitations of other seeds from the same category. See Fig. 3 for a more detailed description of these question types. The results show that even after 8 generations of repetition, imitations can be transcribed into words and matched back to the category of sounds motivating the original imitation."}
gg_match_transcriptions
```

# Discussion {-}

People can be effective at using vocal imitation to represent and communicate about the sounds in their environment, as well as more abstract concepts. Moreover, imitative (or “iconic”) words are found across the spoken languages of the world (Perniss et al. 2010; Dingemanse et al. 2015). However, little is known about the process by which vocal imitations can develop into standardized words. Must new words be deliberately invented as such, or can words form simply by iterated imitations of a sound – even when there is no intention to communicate. To examine this question, we conducted a large-scale, online, iterated vocal imitation experiment.

Our results show that through simple repetition, imitative vocalizations became more word-like – both in form and function. In form, the vocalizations gradually stabilized over generations, becoming more similar from imitation to imitation. They also became increasingly standardized according to the phonology of English, as later generations were more consistently transcribed into English orthography. In function, the imitations became more word-like in their increased effectiveness as category labels. In a learning experiment, naïve participants were faster to learn category labels derived from transcriptions of later-generation imitations than those derived from direct imitations of the environmental sound. This fits with previous research showing that the relatively arbitrary forms that are typical of words (e.g. “dog”) makes them better suited to function as category labels compared to direct auditory cues (e.g. the sound of a bark; Lupyan and Thompson-Schill 2012; Edmiston and Lupyan 2015; Boutonnet and Lupyan, 2015).

However, at the same time as the vocalizations became more word-like, they nevertheless maintained an imitative quality. Interestingly, while after eight generations they could no longer be matched to the particular sound from which they originated, the imitations could still be matched to the general category of the sound. Thus, information that distinguished an imitation from other sound categories was more resilient to transmission decay than exemplar information within a category. Even after the vocalizations were transcribed into English, participants were able to guess their original category from the written “word”.  However, unlike with the vocalizations, participants continued to be more accurate at matching late generation transcriptions back to their particular source sound relative to other exemplars of the category. Thus, with transcriptions, individuating information was retained over generations over and above category information. One possible explanation for this is that by converting the imitations into orthographic representations of phonemes, idiosyncratic features of the sound could become rendered as categorical phonological features. This process could exaggerate the features and facilitate identification of the source.

Our study focused on the process by which words are formed from vocal imitation, and future research remains to determine the full scope of vocal imitation as a source of vocabulary in spoken languages. Although some have estimated the number of imitative words to be small (Crystal, 1987; Newmeyer, 1992), increasing evidence from across disparate languages shows that vocal imitation is, in fact, a widespread source of vocabulary. Cross-linguistic surveys indicate that onomatopoeia – imitative words used to represent sounds – are a universal lexical category found across the world's languages (Dingemanse 2012). Even English, a language that has been characterized as relatively limited in iconic vocabulary (Vigliocco, Perniss, & Vinson 2014), is documented to have hundreds of words for human and animal vocalizations and various kinds of environmental sounds (Rhodes 1994; Sobkowiak 1996). In addition to words that are directly imitative of sounds, many languages also contain semantically broader inventories of ideophones. These words comprise a grammatically and phonologically distinct class of words that are used to express various sensory-rich meanings, such as qualities related to manner of motion, visual properties, textures and touch, inner feelings and cognitive states (Dingemanse, 2012; Nuckolls 1996; Voeltz & Kilian-Hatz 2001). Notably, these words are often recognized by native speakers to bear a degree of resemblance to their meaning, an intuition that is confirmed by experiments with naïve listeners (Dingemanse et al. 2016).

Therefore, if we are to understand the ongoing evolution of spoken languages, it is critical to examine how words are formed from vocal imitation. Here we show that the transition from imitation to word can be a simple process: the mere act of repeated imitation can drive vocalizations to become more word-like in both form and function. Notably, as onomatopoeia and ideophones of natural languages maintain a resemblance to the quality they represent, so did our vocal imitations retain a resemblance to the original sound that inspired them. Altogether, our findings show how words might be created from the repetition of one-shot vocal imitations of an original sound.

# References {-}
