---
title: "Creating words from iterated vocal imitation"
bibliography: telephone.bib
output:
  pdf_document:
    citation_package: natbib
    template: templates/pnas.tex
    keep_tex: true
abstract: |
  We report the results of a large-scale (_N_ = 1571) experiment to investigate whether spoken words can emerge from the process of repeated imitation. Participants played a version of the children’s game “Telephone”. The first generation was asked to imitate recognizable environmental sounds (e.g., glass breaking, water splashing); subsequent generations imitated the imitators for a total of 8 generations. We then examined whether the vocal imitations became more stable and word-like, retained a resemblance to the original sound, and became more suitable as learned category labels. The results showed (1) the imitations became progressively more word-like, (2) even after 8 generations, they could be matched above chance to the category of environmental sound that motivated them, and (3) imitations from later generations were more effective as learned category labels. These results show how repeated imitation can create progressively more word-like forms while retaining a semblance of iconicity.
---

```{r config, include=FALSE}
library(knitr)

opts_chunk$set(
  echo=FALSE,
  message=FALSE,
  warning=FALSE,
  results="hide",
  fig.path="figs/",
  fig.width=7.7,
  fig.height=5.5,
  fig.align="center",
  out.width="7.7cm",  # max is 8.7cm
  dpi=144,
  cache=TRUE,
  cache.path=".cache/",
  autodep=TRUE
)

read_chunk("creating-words.R")
```

```{r setup, include=FALSE, cache=FALSE}
```

People have long pondered the origins of languages, especially the words that compose them. For example, both Plato in his _Cratylus_ dialogue [@Plato:1999uk] and John Locke in his _Essay Concerning Human Understanding_ [@Locke:1948eu] examined the "naturalness" of words--whether they are somehow imitative of their meaning. Some theories of language evolution have hypothesized that vocal imitation played an important role in generating the first words of spoken languages [e.g., @Brown:1955wy; @Donald:2016kd; @Imai:2014dea; @Perlman:2015ip]; early humans may originally have referred to a predatory cat by imitating its roar, or to the discovery of a stream by imitating the sound of rushing water. Such vocal imitation might have served to clarify the referent of a vocalization and eventually establish a mutually understood word. In this study, we investigate the formation of onomatopoeic words -- imitative words that resemble the sounds to which they refer. We ask whether onomatopoeic words can be formed gradually and without instruction simply from repeating the same imitation over generations of speakers.

Onomatopoeic words appear to be a universal lexical category found across the world's languages [@Dingemanse:2012fc]. Languages all have words for animal vocalizations and various environmental sounds that are conventional, but at the same time, exhibit an imitative quality. [@Rhodes:1994au], for example, documented a repertoire of over 100 onomatopoeic words in English, which he notes exist along a continuum from "wild" to "tame". People often use more wild vocal imitations and other sound effects during demonstrative discourse, especially when producing quotations [@Blackwell:2015ix; @Clark:1990cl]. Wild words have a more imitative phonology whereas tame words take on more standard phonology of other words in the language. In some cases, words that begin as wild imitations of sounds become fully lexicalized and integrated into the broader linguistic system, when they behave like more "ordinary" words that can undergo typical morphological processes. Examples are English words like "crack" or the recently adapted "ping".

However, not all researchers agree that vocal imitation has any significant role in language. For instance, [@Pinker:2005cv] suggested that, “Humans are not notably talented at vocal imitation in general, only at imitating speech sounds (and perhaps melodies). For example, most humans lack the ability (found in some birds) to convincingly reproduce environmental sounds … Thus ‘capacity for vocal imitation’ in humans might be better described as a capacity to learn to produce speech.” Nevertheless, experiments show that people can actually be quite effective at using vocal imitation. For example, [@Lemaitre:2014kr] collected imitations and verbal descriptions of various mechanical and synthesized sounds. When participants listened to these and were asked to identify the source, they were more accurate with imitations than descriptions. A subsequent study found that vocal imitations tend to focus on a few salient features of the sound rather than a high fidelity representation, which aids identification of the source [@Lemaitre:2016kz].

Thus humans can be effective at communicating with vocal imitation; it can play an important role in narration and discourse; and it appears to be the basis for substantial inventories of sound-imitative vocabulary across languages. But little is known about the process by which onomatopoeic words like "crack" and "ping" are actually formed and integrated into the vocabulary of a language. A basic question is whether word formation requires deliberation and an intention to create a new word, or whether words can originate from one-shot vocal imitations and repetition. Here we examine whether simple repeated imitations of environmental sounds is sufficient to create more word-like vocalizations, even without an intent to communicate.

To test this, we recruited participants to engage in a large scale online version of the children's game of "Telephone".  In the children’s game, a spoken message is whispered from one person to the next. In our version, the original message was a recording of an environmental sound. The first generation participant imitated the sound, the next generation imitated the previous imitation, and so on. After obtaining these imitations, we investigated how the imitations changed over generations to determine whether they became more word-like. We investigated the acoustic properties of the imitations as well as the orthographic properties once transcribed into English words. We find that by both measures the imitations become more stable through repetition. In addition to stability, we also find that the imitations can still be matched back to the original sounds at above chance levels for many generations. Finally, we measured how quickly the invented words are learned as category labels in a category learning experiment, and find that later generation imitations are easier to learn as category labels.

In Experiment 1 we collected iterated vocal imitations using the transmission chain design depicted in Fig. 1. In the remaining experiments we assessed changes in these imitations over generations. The extent to which each imitation could be matched back to its originating sound was measured in Experiment 2. Experiment 3 involved collecting transcriptions of imitations, and in Experiment 4 these transcriptions were matched back to the original sounds. In Experiment 5 we used transcriptions taken from first and last generation imitations as novel category labels in a simple category learning experiment.

```{r fig1, fig.width=4.6, out.width="4.6cm", fig.align="center", fig.cap="The design of the transmission chain experiment. 16 seed sounds were selected, four in each category of environmental sound. Participants imitated each seed sound, and then the next generation of participants imitated the imitations and so on for 8 generations."}
draw_graphviz("definitions", "wordsintransition")
```

# Results {-}

We begin with a summary of our main results. Measuring the acoustic similarity of repeated imitations revealed that imitations became more similar to one another through repetition. As the imitations were repeated, they gradually lost their resemblence to the source sound. In particular, they lost information that distinguished the source sound from within-category competitors more readily than higher-level category information. This result suggests that through repetition and stabilization the imitations became better abstract category labels by virtue of cuing all category members equally as opposed to highlighting individual category members. We found support for this conclusion in the transcriptions of the imitations. Later generations of imitations were transcribed with better agreement, suggesting that imitations were indeed stabilizing on invented words that were increasingly distinctive and broadly recognizable. Still, these invented words retained some resemblance to the category of environmental sound that motivated them (at least relative to the other categories tested in this experiment). Participants were able to accurately match the transcriptions of final generation imitations in each transmission chain back to the category of environmental sounds that motivated them. Unlike the direct matching of imitations, the extent to which transcriptions were matched to individual source sounds as opposed to categories of sounds did not increase over generations. However, when transcriptions of first and last generation imitations were learned as novel labels of environmental sound categories, last generation transcriptions were easier to learn than those from the first generation. These results describe a process by which an imitation of an environmental sound may transition to a more word-like form through unguided repetition, and suggest that such a transition to more word-like forms might make them more effective as category labels.

## Imitations stabilized over generations

```{r collecting-imitations, include=FALSE}
```

We collected a total of `r n_all_imitations` imitations from `r n_imitators` participants in a study conducted online. `r n_removed` imitations were removed for bad audio quality or violating the rules of the experiment (e.g., saying something in English), leaving `r n_final_imitations` imitations along `r n_branches` contiguous transmission chains for analysis.

Trained research assistants coded these imitations for acoustic similarity using a blinded, pairwise comparison procedure (see Methods). Inter-rater reliability was high, `r report_icc_results(irr_results)`. Imitations from later generations were rated as sounding more similar to one another than imitations from earlier generations, `r lmer_mod_results(similarity_judgments_lmertest_mod, "edge_generation_n")` (Fig. 2). This result suggests that imitations may be stabilizing on particular acoustic forms through repetition.

```{r fig2, fig.height=5.5, fig.cap="Change in acoustic similarity over generations of repetition. Points show mean acoustic similarity ratings for imitations in each category of environmental sounds. The line shows the linear predictions of a hierarchical model with random effects for rater and category, with error bands designating +/- 1 standard error of the model predictions. The results show that acoustic similarity increases over generations, indicating that subsequent imitations become more similar to one another through repetition."}
gg_similarity_judgments
```

## Imitations retained seed category information

```{r matching-imitations, include=FALSE}
```

Were the imitations stabilizing on arbitrary acoustic forms or were they maintaining some aspect of the original environmental sound? To test this, we measured the ability of participants naive to the design of the Telephone game to match each imitation back to its original seed source relative to other seed sounds from either the same category or from different categories (Fig. 3). All 365 imitations were tested in the three conditions depicted in Fig. 3. These conditions differed in the relationship between the imitation and the four seed sounds serving as the choices in the 4 alternative forced choice task.

```{r fig3, fig.width=8, out.width="8.0cm", fig.align="center", fig.cap='Types of 4AFC matching questions depicted in relation to the original set of 16 seed sounds. For each question, participants listened to an imitation (orange dashed circle) and had to guess which of 4 sound choices (green solid circles) they thought the person was trying to imitate. (Top) True seed questions contained the actual sound that generated the imitation in the choices, and the three distractor sounds were sampled from different categories. (Middle) Category match questions also used distractor sounds from different categories but the "correct" sound was not the actual seed, but a different sound within the same category. (Bottom) Specific match questions pitted the actual seed against the other seeds within the same category.'}
grid.arrange(
  q_true_seed,
  q_category_match,
  q_specific_match,
  ncol = 1
)
```

Matching accuracy for all question types started above chance for the first generation of imitations, `r glmer_mod_results(imitation_matches_overall_mod, "(Intercept)", odds = TRUE)`, and decreased steadily over generations, `r glmer_mod_results(imitation_matches_overall_mod, "generation_1")`. We tested whether this increase in matching difficulty was constant across the three types of questions or if some question types became more difficult at later generations than others. The results are shown in Fig. 4. Performance decreased over generations more rapidly for questions requiring a within-category distinction than for between-category questions, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_within")`, suggesting that between-category information was more resistant to loss through transmission. One explanation for this result is that the within-category match questions are simply more difficult because the sounds are more acoustically similar to one another than the between-category questions and therefore performance might be expected to drop off more rapidly with repeated imitations. However, performance also decreased for the easiest type of question where the correct answer was the actual seed generating the imitation (True seed questions; Fig. 3); the advantage of having the true seed among between-category distractors decreased over generations, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_between")`.

```{r fig4, fig.height=5.5, fig.cap='Changes in matching accuracy over generations. Matching accuracy is the ability to guess the sound most likely to have generated the imitation relative to other seed sounds used in the experiment. Performance is separated by question type which describes the relationship between the imitation and the choices in the 4AFC task (see Fig. 3). Lines show predictions from a generalized linear mixed effects model along with +/- 1 standard error of the model predictions. The "category advantage" (category match v. specific match) increased over generations, while the "true seed advantage" (true seed v. category match) decreased. These results suggest that imitions lose within-category information more rapidly than between-category information.'}
gg_match_to_seed
```

These results indicate that as imitations are repeated they lose within-category information more rapidly than between-category information. Later generation imitations were just as likely to be recognized as identifiers of an entire category of environmental sounds as they were of particular sounds within the category.

## Transcription agreement increased over generations

```{r transcriptions, include=FALSE}
```

We next tested whether the imitations became more clearly distinguishable as particular words, as opposed to non-linguistic, i.e., non-English sounds. We had participants transcribe the imitations into English orthography, and then we measured whether transcription agreement increased over generations. We selected the first and final three imitations in each transmission chain to be transcribed. As a control, we also obtained "transcriptions" of the seed sounds themselves. `r n_transcribers` participants generated a total of `r n_transcriptions` or approximately `r n_transcriptions_per_imitation` transcriptions per sound (imitation and seed sounds). Transcriptions containing actual English words and those from participants who failed a catch question were excluded from analysis (`n_transcriptions_dropped`).

```{r table1, results='asis'}
table_caption <- "Examples of invented words."
table_format <- opts_knit$get("out.format")

if(table_format == "markdown") {
  # If it's markdown, you're probably going to tex
  table_format <- "latex"
}

kable(transcription_examples, caption = table_caption,
      format = table_format, booktabs = TRUE)
```

To measure transcription agreement we took the average orthographic distance (`name of algorithm`) between the most frequent transcription and all other transcriptions of a given imitation. Alternative metrics of distance between two pseudo-words resulted in similar results, which are reported in the Supporting Information.

Transcriptions of later generation imitations were more similar to one another in terms of orthographic distance than transcriptions from earlier generations, `r lmer_mod_results(orthographic_distance_lmertest_mod, "message_c")` (Fig. 5). This result supports our hypothesis that unguided repetition drives imitations to become more distinctive as particular English words.

```{r fig5, fig.cap="Average orthographic distance among transcriptions of imitations taken from first and last generations. Each point shows the average orthographic distance between the most frequent transcription and all other transcriptions of a single imitation. Error bars are +/- 1 standard error of the linear mixed effects model predictions. Transcriptions of later generation imitations were more similar to one another than transcriptions of first generation imitations."}
gg_distance
```

## Transcriptions retained seed category information

```{r matching-transcriptions, include=FALSE}
```

We previously demonstrated that people were able to accurately guess the source of an imitation after 8 repetitions, but what about the source of a transcription of an imitation? Do these invented words still resemble the category of sounds that was originally imitated? We tested the top 4 most frequent transcriptions for each imitation in a modified version of the "Guess the Seed" game (see Fig. 3). Participants were given a novel word and had to guess which sound they thought the person who invented the word was talking about.

Participants were able to guess the correct meaning of the transcribed word above chance even after 8 generations of repetition, `r glmer_mod_results(transcription_matches_last_gen_mod, "(Intercept)", odds = TRUE)` (Fig. 6). This was true both for "True seed" questions containing the actual seed generating the transcribed imitation, `r glmer_mod_results(transcription_matches_last_gen_mod_category, "(Intercept)", odds = TRUE)`, and for "Category match" questions where participants had to associate transcriptions with a particular category of environmental sounds, `r glmer_mod_results(transcription_matches_last_gen_mod_exact, "(Intercept)", odds = TRUE)`.

Interestingly, the effect of generation did not vary across these question types, `r glmer_mod_results(acc_mod, "question_c:message_c")`. This indicates that transcriptions of imitations may capture idiosyncratic elements of specific category members more than the imitations themselves. Possible reasons for this asymmetry between imitations and transcriptions are explored in the Discussion.

```{r fig6, fig.cap="Matching accuracy for transcriptions of imitations taken from first and last generations. Bars represent the predicted means taken from the generalized linear mixed effects model with +/- 1 standard error of the model predictions. True seed questions contained transcriptions of the actual seed generating the transcribed word. Category match questions contained transcriptions of imitations of other seeds from the same category. See Fig. 3 for a more detailed description of these question types. The results show that even after 8 generations of repetition, imitations can be transcribed into words and matched back to the category of sounds motivating the original imitation."}
gg_match_transcriptions
```

## Repeated imitations were easier to learn as category labels

```{r category-learning, include=FALSE}
```

Our hypothesis was that repetition of imitations would result in increasingly word-like forms, but what are the consequences of this transition from the perspective of a language user? To examine this question, we tested whether the words created through repetition were easier to learn as category labels.

When participants learned some of the transcriptions as novel category labels for categories of environmental sounds, they were faster when the label came from transcriptions of later generation imitations than from transcriptions of first generation imitations, `r lmer_mod_results(lsn_after_first_block_lmertest_mod, "message_c")` (Fig. 7A).

The effect can be further localized within each block. Comparing RTs on the trials leading up to a block transition and the trials immediately after the block transition revealed a reliable interaction between block transition and the generation of the transcribed label, `r lmer_mod_results(transition_lmertest_mod, "block_transition_c:message_c")` (Fig. 7B). In addition to becoming more stable both in terms of acoustic and orthographic properties, imitations that have been more repeated were also easier to learn as category labels.

```{r fig7, fig.height=11, out.height="11cm", fig.cap="(Top) RTs on correct trials by block, showing faster responses when learning category labels transcribed from last generation imitations. (Bottom) RTs on trials leading up to and immediately following the block transition where new category members are introduced."}
grid.arrange(rt_plot, gg_transition, ncol = 1)
```

# Discussion {-}

Imitative words for sounds, i.e., onomatopoeia, are found across the languages of the world [@Dingemanse:2012fc]. However, little is known about how these words are formed and incorporated into the lexicon of a language. To examine this process, we conducted a large-scale, online, iterated vocal imitation experiment – essentially a version of the children’s game of “Telephone”. The first generation of participants imitated environmental sounds, and then the next generation of participants imitated these imitations, and so on. Our results show that through simple repetition – minus any intention to communicate – the imitations gradually became more word-like. Over generations, they became more stable in sound, and also more easily transcribed into English orthography. However, at the same time, the imitations maintained an onomatopoeic quality: listeners were able to match the vocalizations to both their original sound, and to the sound category, even after eight generations. Even when the imitations were transcribed into English, participants were still able to guess the categorical origin of the word.

The imitations also become more word-like in that they served as more effective category labels. Information that distinguished an imitation from other sound categories was more resilient to transmission decay than exemplar information within a category. Previous research has found that words, as opposed to more veridical cues, make categorization easier [citations]. Similarly, we found that naïve participants were faster to learn category labels derived from transcriptions of later-generation imitations than those derived from direct imitations of the environmental sound. This evidence completes the transition from vocal imitation to abstract word and demonstrates the impact of this transition on communication.

One result that did not fit squarely with imitations becoming more word-like is that with transcriptions, there was no difference over generations between question types. If the results of matching transcriptions back to seed sounds would have perfectly mirrored the results of matching imitations back to seed sounds we would have expected the difference between True seed questions and Category match questions to decrease over generations. Instead we found a main effect of question type. Although participants were able to match transcriptions to categories of sounds even after 8 generations of repetition, it was still easier for them to match a transcription to the actual seed that generated the transcription.

Our study focused on the formation of onomatopoeia -- sound-imitative words -- but in addition to onomatopoeia, many languages have semantically rich systems of ideophones. These words comprise a grammatically and phonologically distinct class of words that are used to express a variety of sensory-rich meanings [@Dingemanse:2012fc; @Voeltz:2001vv]. Notably, these words are often recognized by native speakers to be imitative of their meaning to some degree. For example, in Japanese, the word ‘koron’ -- with a voiceless [k] -- refers to a light object rolling once, the reduplicated ‘korokoro’ to a light object rolling repeatedly, and ‘gorogoro’ -- with a voiced [g] -- to a heavy object rolling repeatedly [@Imai:2014dea]. The iconicity of ideophones was verified by results showing that naïve listeners were able to guess the meanings of words sampled from five different languages [@Dingemanse:2016vd]. Although words for sounds were guessed more accurately than the rest, listeners were better than chance at guessing the synonyms of ideophones that expressed meanings from all five semantic categories tested, which also included color/visual, motion, shape, and texture.  In addition, laboratory experiments show that people are able to generate imitative vocalizations for a variety of non-sound concepts, and that these are also understandable to naïve listeners [@Perlman:2015ip]. Thus vocal imitation has the potential to play a role in word formation that extends beyond just the imitation of sounds.

Our findings from an online game of Telephone suggest that the formation of words from vocal imitation can be a simple process. The results show how repeated imitation can create progressively more word-like forms while retaining a resemblance to the original sound that motivated it. This raises the possibility that onomatopoeic words can be created from the repetition of one-shot vocal imitations of an original sound.

# References {-}
