---
title: "The emergence of words from vocal imitations"
bibliography: telephone.bib
output:
  pdf_document:
    citation_package: natbib
    template: templates/pnas.tex
    keep_tex: true
abstract: |
  We investigated how conventional spoken words might emerge from imitations of environmental sounds. Participants played a version of the children’s game “Telephone”. The first generation of participants imitated recognizable environmental sounds (e.g., glass breaking, water splashing). Subsequent generations imitated the imitations of the prior generation for a maximum of 8 generations. The results showed that the imitations became more stable and word-like, and more easily learnable as category labels. At the same time, even after 8 generations, both spoken imitations and their written transcriptions could be matched above chance to the category of environmental sound that motivated them. These results show how repeated imitation can create progressively more word-like forms that continue to retain a resemblance to the original sound that motivated them. The results speak to the possible role of human vocal imitation in explaining the origins of spoken words.
nocite: |
  @PCT:1.1
---

People have long pondered the origins of languages, especially the words that compose them. For example, both Plato in his _Cratylus_ dialogue [@Plato:1999uk] and John Locke in his _Essay Concerning Human Understanding_ [@Locke:1948eu] examined the "naturalness" of words---whether they are somehow imitative of their meaning. Here we investigated whether new words can be formed from the repetition of non-verbal vocal imitations. Does the repetition of imitations over generations of speakers gradually give rise to novel word forms? In what ways do these words resemble the original sounds that motivated them? We report a large-scale experiment (_N_=1571) investigating how new words can form---gradually and without instruction---simply by repeating imitations of environmental sounds.

The importance of imitation and depiction in the origin of signs is clearly observable in the origin of words in signed languages [@Klima:1980si; @GoldinMeadow:2016bw; @Kendon:2014eg], but in considering the idea that vocal imitation may be key to understanding the origin of spoken words, many have argued that the human capacity for vocal imitation is far too limited to play a significant role [@Arbib:2012htb; @Tomasello:2010or; @Armstrong:2007go; @Corballis:2003ha; @Hockett:1978se; @Hewes:1973vr]. For example, Pinker and Jackendoff [@Pinker:2005cv] argued that, “most humans lack the ability… to convincingly reproduce environmental sounds… Thus ‘capacity for vocal imitation’ in humans might be better described as a capacity to learn to produce speech” (p. 209). Consequently, it is still widely assumed that vocal imitation---or more broadly, the use of any sort of resemblance between form and meaning---cannot be important to understanding the origin of spoken words.

Although most words of contemporary spoken languages are not clearly imitative in origin, there has been a growing recognition of the importance of imitative words in spoken languages [@Dingemanse:2015cu; @Perniss:2010fb] and the frequent use of vocal imitation and depiction in spoken discourse [@Clark:1990cl; @Lewis:2009wz]. This has led some to argue for the importance of imitation for understanding the origin of spoken words [e.g., @Brown:1955wy; @Donald:2016kd; @Imai:2014dea; @Perlman:2015ip; @Dingemanse:2014gj]. In addition, experiments show that counter to previous assumptions, people are highly effective at using vocal imitations in reference---in some cases, even more effective than with conventional words [@Lemaitre:2014kr]. Recent work has also shown that people are able to create novel imitative vocalizations for more abstract meanings (e.g. ‘slow’, ‘rough’, ‘good’, ‘many’) that are understandable to naïve listeners [@Perlman:2015ip]. The effectiveness of these imitations arises not because people can mimic environmental sounds with high fidelity, but because they are able to produce imitations that capture the salient features of sounds in ways that are understandable to listeners [@Lemaitre:2016kz]. Similarly, the features of onomatopoeic words might highlight distinctive aspects of the sounds they represent. For example, the initial voiced, plosive /b/ in “boom” represents an abrupt, loud onset, the back vowel /u/ a low pitch, and the nasalized /m/ a slow, muffled decay [@Rhodes:1994au].

Thus, converging evidence suggests that people can use vocal imitation as an effective means of communication. But can vocal imitations give rise to words that can be integrated into the vocabulary of a language? And if so, what is required for this to happen? What happens to a vocal imitation in the course of it being turned into a word? To answer these questions, we recruited participants to play an online version of the children's game of "Telephone". In the children’s game, a spoken message is whispered from one person to the next. In our version, the original message or "seed sound" was a recording of an environmental sound. The initial group (first generation) of participants imitated these seed sounds, the next generation imitated the previous imitators, and so on for up to 8 generations (Fig. 1).

In subsequent experiments, we systematically answered the following questions about the form of the vocalizations and their potential to function as words. First, does iterated imitation drive the vocalizations to stabilize in form and become more word-like? Second, do the imitations become more suitable as labels for the category of sounds that motivated them? For example, does the imitation of a particular water-splashing sound become, over time, a better label for the more general category of water-splashing sounds? Third, do the imitations retain a resemblance to the original environmental sounds that inspired them? If so, it should be possible for naïve participants to match the emergent words back to the seed sounds that were originally imitated.

```{r config, include=FALSE}
library(knitr)

opts_chunk$set(
  echo=FALSE,
  message=FALSE,
  warning=FALSE,
  results="hide",
  fig.path="figs/",
  fig.align="center",
  fig.width=7.7,
  fig.height=5.5,
  out.width="7.7cm",  # max is 8.7cm
  out.height="5.5cm",
  dpi=144,
  cache=TRUE,
  cache.path=".cache/",
  autodep=TRUE
)

read_chunk("R/0-setup.R")
read_chunk("R/1-telephone-game.R")
read_chunk("R/2-stability.R")
read_chunk("R/3-learning.R")
read_chunk("R/4-matching.R")
```

```{r setup, include=FALSE, cache=FALSE}
```

```{r dendrogram, include=FALSE}
```

```{r fig1, fig.width=8.7, out.width="8.7cm", fig.align="center", fig.cap="The design of the transmission chain experiment. Seed sounds (16) were sampled from four categories of environmental sounds: glass, tear, water, zipper. Participants imitated each seed sound, and then the next generation of participants imitated the imitations and so on for up to 8 generations. Chains are unbalanced due to random assignment and the exclusion of some low quality recordings."}
gg_dendrogram
```

# Results {-}

We begin with a summary of our main results: (1) Imitations of environmental sounds became more stable over the course of being repeatedly imitated as revealed by increasing acoustic similarity along individual transmission chains. In addition, later generations of imitations had higher levels of agreement when transcribed into English orthography further suggesting an increase in stability and word-likeness. (2) When transcriptions of first and last generation imitations were learned as novel labels for categories of environmental sounds, last generation transcriptions were learned faster and generalized to new category members more easily than transcriptions of first generation imitations, suggesting that repeating imitations caused the forms to become more suitable as category labels. (3) Even as the imitations became more word-like, they also retained a resemblance to the original category of environmental sound that motivated them, as measured by the ability of naïve listeners to match both the auditory imitations and their written transcriptions to the correct category of environmental sounds even after 8 generations of repetition.

## Iterated imitations became more stable and word-like

```{r stability, include=FALSE}
```

The final set of vocal imitations included `r n_final_imitations` imitations along `r n_branches` contiguous transmission chains from `r n_imitators` participants (Fig. 1; see Methods). Research assistants (_N_=5) rated the acoustic similarity of pairs of imitations while blind to all conditions and hypotheses (see Methods). We also conducted automated analyses of acoustic similarity using Mel Frequency Cepstral Coefficients (MFCCs) as a measure of acoustic distance. These results are reported in the Supporting Information (Fig. S5). Acoustic similarity ratings were fit with a hierarchical linear model[^df] predicting similarity from generation with random effects for rater and for category. Imitations from later generations were rated as sounding more similar to one another than imitations from earlier generations, `r lmer_mod_results(similarity_judgments_lmertest_mod, "edge_generation_n")` (Fig. 2A). This result suggests that imitations became more stable (i.e., easier to imitate with high fidelity) with each generation.

[^df]: Degrees of freedom and corresponding significance tests for hierarchical linear models were estimated using the Satterthwaite approximation [@lmerTest:2016].

```{r fig2, fig.height=12, out.height="12cm", fig.cap="Stabilization of imitations through repetition. A. Change in perception of acoustic similarity over generations of repetition. Means and hierarchical linear model predictions with ±1 SE are shown. Acoustic similarity increased over generations, indicating that repetition made the vocalizations easier to imitate with high fidelity. B. Transcription similarity for first and last generation imitations. Mean orthographic distance between the most frequent transcription and all other transcriptions of a given imitation are shown, with error bars as ±1 SE of the hierarchical linear model predictions. Transcriptions of later generation imitations were more similar to one another than transcriptions of first generation imitations."}
grid.arrange(
  gg_similarity_judgments,
  gg_distance,
  ncol = 1
)
```

As an additional test of stabilization, we had English-speaking participants transcribe a sample of first and last generation imitations into English orthography, and then measured whether transcription similarity (spelling agreement) increased over generations. We collected a total of `r n_transcriptions` transcriptions---approximately `r n_transcriptions_per_imitation` transcriptions per sound (see Methods). Some examples of the transcriptions are presented in Table 1.

```{r table1, results='asis'}
table_caption <- "Examples of invented words."
table_format <- opts_knit$get("out.format")

if(table_format == "markdown") {
  # If it's markdown, you're probably going to tex eventually
  table_format <- "latex"
}

kable(transcription_examples, caption = table_caption,
      format = table_format, booktabs = TRUE)
```

To measure the similarity among transcriptions of a given imitation, we calculated the orthographic distance between the most frequent transcription and all other transcriptions. The orthographic distance measure was a ratio based on longest contiguous matching subsequences between pairs of transcriptions. A hierarchical linear model predicting orthographic distance from the generation of the imitation (First generation, Last generation) with random effects for transmission chains nested within categories of environmental sounds revealed that transcriptions of last generation imitations were more similar to one another than transcriptions from first generation imitations, `r lmer_mod_results(orthographic_distance_lmertest_mod, "message_c")` (Fig. 2B). The same result is reached through alternative measures of orthographic distance such as exact string matching and length of longest substring match, and when excluding imitations for which all transcriptions were unique in which case there was no most frequent transcription (Fig. S9). These results support our hypothesis that unguided repetition drives imitations to stabilize on particular words.

## Iterated imitations made for better category labels

```{r learning, include=FALSE}
```

One consequence of imitations becoming more word-like is that they may make for better category labels. For example, an imitation from a later generation, by virtue of having a more word-like form, may be easier to learn as a label for the category of sounds that motivated it than an earlier imitation, which may be more closely yoked to an individual seed sound. To the extent that repeating imitations abstracts away the idiosyncrasies of a particular category member, it may also be easier to generalize to new category members. We tested these predictions using a category learning task wherein participants had to learn novel labels for categories of environmental sounds. Unbeknownst to the participants, the novel labels they learned were transcriptions generated either from first or last generation imitations. The procedure for selecting otherwise-equal transcriptions is detailed in the Supporting Information. Here we focus on the consequences of learning either first or last generation transcriptions in the category learning experiment.

At the beginning of the experiment, where participants had to learn through trial-and-error which labels were associated with which sounds, participants learning transcriptions of first or last generation imitations did not differ in overall accuracy, `r glmer_mod_results(lsn_first_block_error_mod, "message_c", p_value_only = TRUE)`, or reaction time, `r lmer_mod_results(lsn_first_block_rt_mod, "message_c", p_value_only = TRUE)`. After this initial learning phase (i.e. after the first block of trials), accuracy performance quickly reached ceiling (Fig. S12) and did not differ between groups `r glmer_mod_results(lsn_after_first_block_error_mod, "message_c", p_value_only = TRUE)`. However, participants learning last generation transcriptions responded more quickly in subsequent blocks than participants learning first generation transcriptions, `r lmer_mod_results(lsn_after_first_block_lmertest_mod, "message_c")` (Fig. 3A). These faster responses suggest that, in addition to becoming more stable both in terms of acoustic and orthographic properties, repeating imitations makes them easier to learn as category labels. Given how quickly accuracy performance reached ceiling, further investigation with a more difficult category learning experiment is warranted (e.g., more than four categories and 16 exemplars).

Next, we examined whether transcriptions from last generation imitations were easier to generalize to novel sounds. To test this hypothesis, we compared RTs on trials immediately prior to the introduction of novel sounds (new category members) and the first trials after the block transition (±6 trials). The results revealed a reliable interaction between the generation of the transcribed imitation and the block transition, `r lmer_mod_results(transition_lmertest_mod, "block_transition_c:message_c")` (Fig. 3B). This result suggests that transcriptions from later generation imitations were easier to generalize to new category members. 

```{r fig3, fig.height=10, out.height="10cm", fig.cap="Learning transcriptions of imitations as category labels. A. Participants achieved faster RTs in matching transcribed labels to environmental sounds for labels transcribed from later compared to earlier generation imitations. B. There was a generalization cost for first-generation labels, but not last generation labels."}
grid.arrange(rt_plot, gg_transition, ncol = 1)
```

## Iterated imitations retained resemblance to original sounds

```{r matching, include=FALSE}
```

As the imitations became more word-like, were they stabilizing on arbitrary acoustic and orthographic forms, or did they maintain some resemblance to the original environmental sound that motivated them? To test this, we measured the ability of participants naïve to the design of the experiment to match imitations back to their original source relative to other seed sounds from either the same category or from different categories (Fig. 4A). All 365 imitations were tested in the three question types depicted in Fig. 4A. These questions differed in the relationship between the imitation and the four seed sounds provided as the choices in the question. Responses were fit by a hierarchical generalized linear model predicting match accuracy as different from chance (25%) based on the type of question being answered (True seed, Category match, Specific match) and the generation of the imitation.

```{r fig4, fig.width=8.7, out.width="8.7cm", fig.height=17, out.height="17cm", fig.align="center", fig.cap='A. Three types of matching questions used to assess the resemblance between the imitation (and transcriptions of imitations) and the original seed sounds. For each question, participants listened to an imitation (dashed circles) and had to guess which of 4 sound choices (solid circles) they thought the person was trying to imitate. True seed questions contained the actual sound that generated the imitation as one of the choices (correct response). The remaining sounds were sampled from different categories. Category match questions replaced the original seed sound with another sound from the same category. Specific match questions pitted the actual seed against the other seeds within the same category. B. Change in matching accuracy over generations of imitations, shown as predictions of the generalized linear models with ±1 SE of the model predictions. The "category advantage" (Category match vs. Specific match) increased over generations, while the "true seed advantage" (True seed v. Category match) decreased (see main text), suggesting that imitations lose within-category information more rapidly than between-category information. C. Change in matching accuracy over generations of imitations transcribed into English-sounding words. Imitations and transcriptions of imitations could still be matched back to the category of sound that motivated the original imitation even after 8 generations. Match accuracy for imitations is shown for comparison.'}
q_true_seed <- read_graphviz("true-seed", "wordsintransition")
q_category_match <- read_graphviz("category-match", "wordsintransition")
q_specific_match <- read_graphviz("specific-match", "wordsintransition")

grid.arrange(
  arrangeGrob(q_true_seed, q_category_match, q_specific_match, ncol = 1,
              top = textGrob('A. Types of "Guess the seed" matching questions',
                             hjust = 0.7, gp = gpar(fontsize = 16))),
  gg_match_to_seed,
  gg_match_transcriptions,
  ncol = 1,
  heights = c(0.4, 0.3, 0.3)
)
```

Matching accuracy for all question types was above chance for the first generation of imitations, `r glmer_mod_results(imitation_matches_overall_mod, "(Intercept)", odds = TRUE)`, and decreased steadily over generations, `r glmer_mod_results(imitation_matches_overall_mod, "generation_1")`. We tested whether this increase in difficulty was constant across the three types of questions or if some question types became more difficult than others. The results are shown in Fig. 4B. Performance decreased over generations more rapidly for questions requiring a within-category distinction than for between-category questions, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_within")`, suggesting that between-category information was more resistant to loss through repeated imitation. An alternative explanation for this result is that the within-category match questions are simply more difficult[^below] because the sounds provided as choices are more acoustically similar to one another than the between-category questions, and therefore, performance might be expected to drop off more rapidly with repeated imitations. However, performance also decreased for the easiest type of question where the correct answer was the actual seed generating the imitation (True seed questions; see Fig. 4A); the advantage of having the true seed among between-category distractors decreased over generations, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_between")`. The observed increase in the "category advantage" (i.e., the advantage of having between-category distractors) combined with a decrease in the "true seed advantage" (the advantage of having the actual seed among the choices), shows that the changes induced by repeated imitation caused the imitations to lose some of properties that linked the earlier imitations to the specific sound that motivated them, while nevertheless preserving a more abstract category-based resemblance.

[^below]: We observed that performance on some Specific match questions dropped below chance for later generations indicating participants had an apparent aversion to the nominally correct answer. Additional analyses showed that participants were not converging on a single incorrect response. The reason for this pattern is at present unclear. Removing these trials from the analysis does not substantively change the conclusions.

We next tested whether it was possible to match the written transcriptions of the auditory sounds back to the original environmental sounds. Participants were given a novel word (the most frequent transcriptions of first and last generation imitations) and had to guess the sound that was represented by the invented word. The distractors for all questions were between-category, i.e. true seed and category match. Specific match questions were omitted.

Remarkably, participants were able to guess the correct meaning of a word that was transcribed from an imitation that had been repeated up to 8 times, `r glmer_mod_results(transcription_matches_last_gen_mod, "(Intercept)", odds = TRUE)` (Fig. 4C). This was true for True seed questions containing the actual seed generating the transcribed imitation, `r glmer_mod_results(transcription_matches_last_gen_mod_category, "(Intercept)")`, and for Category match questions where participants had to associate transcriptions with a particular category of environmental sounds, `r glmer_mod_results(transcription_matches_last_gen_mod_exact, "(Intercept)")`. The effect of generation did not vary across these question types, `r glmer_mod_results(acc_mod, "question_c:message_c")`. Possible reasons for this difference between imitations and their transcriptions are explored in the Supporting Information.

In sum, our results show how unguided repetition causes initial imitations of environmental sounds transition to more word-like forms. They suggest that in the course of this transition, the imitations become more categorical and more effective as learned category labels all while retaining some resemblance to the environmental sounds that motivated them.

# Discussion {-}

Imitative (or "iconic") words are found across the spoken languages of the world [@Dingemanse:2015cu; @Imai:2014dea; @Perniss:2010fb]. Counter to past assumptions about the limitations of human vocal imitation, people are surprisingly effective at using vocal imitation to represent and communicate about the sounds in their environment [@Lemaitre:2016kz] and more abstract meanings [@Perlman:2015ip], making the hypothesis that early spoken words originated from imitations a plausible one. We examined whether simply repeating an imitation of an environmental sound---with no intention to create a new word or even to communicate---produces more word-like forms.

Our results show that through simple repetition, imitative vocalizations became more word-like both in form and function. In form, the vocalizations gradually stabilized over generations, becoming more similar from imitation to imitation. They also became increasingly standardized in accordance with English orthography, as later generations were more consistently transcribed into English words. In function, the increasingly word-like forms became more effective as category labels. In a category learning experiment, naïve participants were faster to learn category labels derived from transcriptions of later-generation imitations than those derived from direct imitations of the environmental sound. This fits with previous research showing that the relatively arbitrary forms that are typical of words (e.g. “dog”) makes them better suited to function as category labels compared to direct auditory cues [e.g. the sound of a dog bark; @Lupyan:2012cp; @Edmiston:2015he; @Boutonnet:2015fz].

Evan as the vocalizations became more word-like, they nevertheless maintained an imitative quality. After eight generations they could no longer be matched to the particular sound from which they originated any more accurately than they could be matched to the general category of environmental sound. Thus, information that distinguished an imitation from other sound categories was more resilient to transmission decay than exemplar information within a category. Remarkably, even after the vocalizations were transcribed into English orthography, participants were able to guess their original sound category from the written “word”.  In contrast to the vocalizations, participants continued to be more accurate at matching late generation transcriptions back to their particular source sound relative to other exemplars from the same category.

Although the number of imitative words in contemporary languages may appear to be very small [@Crystal:1987en; @Newmeyer:1992we], increasing evidence from disparate languages shows that vocal imitation is, in fact, a widespread source of vocabulary. Cross-linguistic surveys indicate that onomatopoeia---imitative words used to represent sounds---are a universal lexical category found across the world's languages [@Dingemanse:2012fc]. Even English, a language that has been characterized as relatively limited in iconic vocabulary [@Vigliocco:2014fc], is documented to have hundreds of clearly imitative words including words for human and animal vocalizations as well as various types of environmental sounds [@Rhodes:1994au; @Sobkowiak:1990ph]. Besides words that are directly imitative of sounds---the focus of the present study---many languages contain semantically broader inventories of ideophones. These words comprise a grammatically and phonologically distinct class of words that are used to express various sensory-rich meanings, such as qualities related to manner of motion, visual properties, textures and touch, inner feelings and cognitive states [@Dingemanse:2012fc; @Nuckolls:1999ca; @Voeltz:2001vv]. As with onomatopoeia, ideophones are often recognized by naïve speakers as bearing a degree of resemblance to their meaning [@Dingemanse:2016vd].

Our study focused on imitations of environmental sounds and more work remains to be done to determine the extent to which vocal imitation can ground de novo vocabulary creation in other semantic domains [e.g., @Perlman:2015ip; @Lupyan:2015vic]. What the present results make clear is that the transition from imitation to word can be a rapid and simple process: the mere act of repeated imitation can drive vocalizations to become more word-like in both form and function. Notably, just as onomatopoeia and ideophones of natural languages maintain a resemblance to the quality they represent, the present vocal imitations transitioned to words while retaining a resemblance to the original sound that motivated them. 

# References {-}
