---
title: "Creating words from iterated vocal imitation"
bibliography: telephone.bib
output:
  pdf_document:
    citation_package: natbib
    template: templates/pnas.tex
    keep_tex: true
abstract: |
  We report the results of a large-scale (_N_ = 1571) experiment to investigate
  whether spoken words can emerge from the process of repeated imitation.
  Participants played a version of the children’s game “Telephone”. The first
  generation was asked to imitate recognizable environmental sounds (e.g., glass
  breaking, water splashing); subsequent generations imitated the imitators for
  a total of 8 generations. We then examined whether the vocal imitations became
  more stable and word-like, retained a resemblance to the original sound, and
  became more suitable as learned category labels. The results showed (1) the
  imitations became progressively more word-like, (2) even after 8 generations,
  they could be matched above chance to the category of environmental sound that motivated
  them, and (3) imitations from later generations were more effective as learned
  category labels. These results show how repeated imitation can create
  progressively more word-like forms while retaining a semblance of iconicity.
---

```{r config, include=FALSE}
library(knitr)

opts_chunk$set(
  echo=FALSE,
  message=FALSE,
  warning=FALSE,
  results="hide",
  fig.path="figs/",
  fig.width=8.7,
  fig.height=6,
  fig.align="center",
  out.width="8.7cm",  # max is 8.7cm
  dpi=144,
  cache=TRUE,
  cache.path=".cache/"
)

read_chunk("creating-words.R")
```

```{r setup, include=FALSE}
```

People have long pondered the origins of languages, especially the
words that compose them. For example, both Plato in his _Cratylus_ dialogue
[@Plato:1999uk] and John Locke in his _Essay Concerning Human Understanding_
[@Locke:1948eu] examined the "naturalness" of words--whether they are somehow
imitative of their meaning. Some theories of language evolution have
hypothesized that vocal imitation played an important role in generating the
first words of spoken languages [e.g., @Brown:1955wy; @Donald:2016kd;
@Imai:2014dea; @Perlman:2015ip]; early humans may originally
have referred to a predatory cat by imitating its roar, or to the discovery of a
stream by imitating the sound of rushing water. Such vocal imitation might have
served to clarify the referent of a vocalization and eventually establish a
mutually understood word. In this study, we investigate the formation of
onomatopoeic words--imitative words that resemble the sounds to which they
refer. We ask whether onomatopoeic words can be formed gradually and without
instruction simply from repeating the same imitation over generations of
speakers.

Onomatopoeic words appear to be a universal lexical category found across the
world's languages [@Dingemanse:2012fc]. Languages all have conventional words
for animal vocalizations and various environmental sounds. [@Rhodes:1994au], for
example, documented a repertoire of over 100 onomatopoeic words in English,
which he notes exist along a continuum from "wild" to "tame". People often use
more wild vocal imitations and other sound effects during demonstrative
discourse, especially when producing quotations [@Blackwell:2015ix;
@Clark:1990cl]. Wild words have a more imitative phonology whereas tame words
take on more standard phonology of other English words. In some cases, words
that begin as wild imitations of sounds become fully lexicalized and integrated
into the broader linguistic system, when they behave like more "ordinary" words
that can undergo typical morphological processes. Examples are English words
like "crack" or the recently adapted "ping".

However, not all researchers agree that vocal imitation has any significant
role in language. For instance, [@Pinker:2005cv] suggested that, “Humans are
not notably talented at vocal imitation in general, only at imitating speech
sounds (and perhaps melodies). For example, most humans lack the ability (found
in some birds) to convincingly reproduce environmental sounds … Thus ‘capacity
for vocal imitation’ in humans might be better described as a capacity to learn
to produce speech.” Nevertheless, experiments show that people can actually be
quite effective at using vocal imitation. For example, [@Lemaitre:2014kr]
collected imitations and verbal descriptions of various mechanical and
synthesized sounds. When participants listened to these and were asked to
identify the source, they were more accurate with imitations than descriptions.
A subsequent study found that vocal imitations tend to focus on a few salient
features of the sound rather than a high fidelity representation, which aids
identification of the source [@Lemaitre:2016kz].

Thus humans can be effective at communicating with vocal imitation, it can play
an important role in narration and discourse, and it appears to be the basis for
substantial inventories of sound-imitative vocabulary across languages. But the
process by which onomatopoeic words like "meow", "ping" and "buzz" emerge from
vocal imitations has yet to be observed. Here we examine whether simple repeated
imitations of environmental sounds become more word-like even in the absence of
explicit communication intent or the intent to create a word-like token.
Alternatively, repeating imitations might never stabilize on a particular
word form, or the limited fidelity of human vocal imitation may simply restrict
the formation of stable words through iterated imitations.

To test this, we recruited participants to engage in a large scale online
version of the children's game of "Telephone" in which an acoustic message is
passed from one person to the next. After obtaining these imitations, we
investigated how the imitations changed over generations to determine whether
they became more word-like. We investigated the acoustic properties of the
imitations as well as the orthographic properties once transcribed into English
words. We find that by both measures the imitations become more stable through
repetition. In addition to stability, we also find that the imitations can still
be matched back to the original sounds at above chance levels for many
generations. Finally, we measured how quickly the invented words are learned as
category labels in a category learning experiment, and find that later
generation imitations are easier to learn as category labels.

In Experiment 1 we collected iterated vocal imitations using the transmission chain design depicted in Fig. 1. In the remaining experiments we assessed changes in these imitations over generations. The extent to which each imitation could be matched back to its originating sound was measured in Experiment 2. Experiment 3 involved collecting transcriptions of imitations, and in Experiment 4 these transcriptions were matched back to the original sounds. In Experiment 5 we used transcriptions taken from first and last generation imitations as novel category labels in a simple category learning experiment.

```{r fig1, fig.width=4.6, out.width="4.6cm", fig.align="center", fig.cap="The design of the transmission chain experiment. 16 seed sounds were selected, four in each category of environmental sound. Participants imitated each seed sound, and then the next generation of participants imitated the imitations and so on for 8 generations."}
draw_graphviz("definitions", "wordsintransition")
```

# Results {-}

A summary of our results is as follows. Measuring the acoustic similarity of subsequent imitations revealed that imitations became more similar to one another over generations of repetition. The more times imitations were repeated, the harder they were to match back to the original sounds, and within-category information was lost more readily than between-category information. Similar results are found when analyzing transcriptions of the imitations. Transcription agreement increases over generations of imitation, and transcriptions of even last generation imitations can be matched back to the category of environmental sounds. When transcriptions of first and last generation imitations are learned as novel category labels, last generation transcriptions were easier to learn than first generation transcriptions. These results describe a process by which an imitation of an environmental sound may transition to a more word-like and categorical form through unguided repetition.

## Imitations stabilized over generations

```{r collecting-imitations, include=FALSE}
```

We collected a total of `r n_all_imitations` imitations from `r n_imitators` in an online study. `r n_removed` imitations were removed for bad audio quality or violating the rules of the experiment (e.g., saying something in English), leaving `r n_final_imitations` imitations along `r n_branches` contiguous transmission chains for analysis. Trained research assistants coded these imitations for acoustic similarity using a blinded, pairwise comparison procedure (see Methods; `intercode reliability`). Imitations from later generations were rated as being more similar to one another than imitations from earlier generations, `r lmer_mod_results(similarity_judgments_lmertest_mod, "edge_generation_n")` (Fig. 2).

```{r fig2, fig.cap="Increase in acoustic similarity over generations. Points depict mean acoustic similarity ratings for imitations in each category of environmental sounds. The predictions of the linear mixed effects model with random effects for rater and category are shown, with error bands denoting +/- 1 standard error of the model predictions."}
gg_similarity_judgments
```

## Imitations retained category information

```{r matching-imitations, include=FALSE}
```

Were the imitations stabilizing on arbitrary acoustic forms or were they capturing some aspect of the original environmental sound? To test this, we measured the ability of naive participants to match each imitation back to its original source relative to other seed sounds from the same category or from different categories. All `r n_final_imitations` imitations were tested in the three conditions depicted in Fig. 3. These conditions differed in the relationship between the imitation and the four seed sounds serving as the options in the 4AFC task.

```{r fig3, fig.width=8, out.width="8.0cm", fig.align="center", fig.cap='Types of matching questions in the "Guess the seed" game. For each question, participants listened to an imitation (orange circle) and had to guess which of 4 sound choices (green circles) they thought the person was trying to imitate. White circles depict the original set of 16 seed sounds (not shown to participants). (Top) True seed questions contained the actual seed that generated the imitation in the choices, and the distractor seeds were sampled from different categories. (Middle) Category match questions also used distractor sounds from different categories but the correct seed was not the actual seed, but a different sound within the same category. (Bottom) Specific match questions pitted the actual seed against the other seeds within the same category.'}
grid.arrange(
  q_true_seed,
  q_category_match,
  q_specific_match,
  ncol = 1
)
```

Matching accuracy for all question types started above chance for the first generation of imitations, `r glmer_mod_results(imitation_matches_overall_mod, "(Intercept)", odds = TRUE)`, and decreased steadily over generations, `r glmer_mod_results(imitation_matches_overall_mod, "generation_1")`. We tested whether this increase in matching difficulty was constant across the three types of questions or if some question types became more difficult at later generations.

The results are shown in Fig. 4. Performance decreased over generations more rapidly for specific match questions than for category match questions, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_between")`, suggesting that between-category information was more resistant to loss through transmission. One explanation for this result is that the specific match questions are simply more difficult than the category match questions and therefore performance might be expected to drop off more rapidly. However, performance also decreased for the easiest type of question where the correct answer was the actual seed generating the imitation; the advantage of having the true seed among between-category distractors decreased over generations, `r glmer_mod_results(imitation_matches_mod, "generation_1:same_v_within")`. These results indicate that as imitations were repeated they lost within-category information more rapidly than between-category information. Later generation imitations were just as likely to be recognized as identifiers of a particular category of environmental sounds as they were of particular exemplar sounds.

```{r fig4, fig.cap='Accuracy in matching imitations back to seed sounds. Performance is separated by question type based on the relationship between the imitation and the options in the question (see Fig. 3). Lines depict predictions from the generalized linear mixed effects model along with +/- 1 standard error of the model predictions. The "category advantage" (category match v. specific match) increases over generations, while the "true seed advantage" (true seed v. category match) decreases. These results suggest that imitions lose within-category information more rapidly than between-category information.'}
gg_match_to_seed
```

## Transcription agreement increased over generations

```{r transcriptions, include=FALSE}
```

We next tested whether the imitations were stabilizing on particular words as opposed to non-linguistic, i.e., non-English sounds by having participants transcribe the imitations and measuring whether transcription agreement increased over generations. We selected the first and final three imitations in each transmission chain to be transcribed into English orthography. A total of `r n_transcriptions` or `r n_transcriptions_per_imitation` transcriptions per imitation were collected. All transcriptions containing actual English words were excluded from analysis (`n_transcriptions_dropped`). Transcriptions from later generation imitations were more similar to one another in terms of orthographic distance than transcriptions from earlier generations, `r lmer_mod_results(orthographic_distance_lmertest_mod, "message_c")` (Fig. 5).

```{r fig5, fig.cap="Average orthographic distance among transcriptions of imitations taken from first and last generations. Each point shows the average orthographic distance between the most frequent transcription and all other transcriptions of a single imitation. Error bars are +/- 1 standard error of the linear mixed effects model predictions. Transcriptions of later generation imitations were more similar to one another than transcriptions of first generation imitations."}
gg_distance
```

## Transcriptions retained association with seed category

```{r matching-transcriptions, include=FALSE}
```

Participants were able to associate imitations after 8 repetitions with the sound that originally motivated them, but what about the transcriptions of these imitations? Do these invented words still resemble the category of sounds that was originally imitated? We tested the top 4 most frequent transcriptions for each imitation in a modified version of the "Guess the Seed" game (see Fig. 3). Participants were able to guess the correct meaning of the transcribed word above chance even after 8 generations of repetition, `r glmer_mod_results(transcription_matches_last_gen_mod, "(Intercept)", odds = TRUE)` (Fig. 6). This was true both for "true seed" questions containing the actual seed generating the transcribed imitation, `r glmer_mod_results(transcription_matches_last_gen_mod_category, "(Intercept)", odds = TRUE)`, and for category match questions where participants had to associate transcriptions with a particular category of environmental sounds, `r glmer_mod_results(transcription_matches_last_gen_mod_exact, "(Intercept)", odds = TRUE)`. The effect of generation did not vary across these question types, `r glmer_mod_results(acc_mod, "question_c:message_c")`.

```{r fig6, fig.cap="Matching accuracy for transcriptions of imitations taken from first and last generations. True seed questions contained transcriptions of the actual seed generating the transcribed word. Category match questions contained transcriptions of imitations of other seeds from the same category."}
gg_match_transcriptions
```

## Repeated imitations were easier to learn as category labels

```{r category-learning, include=FALSE}
```

When participants learned some of the transcriptions as novel category labels for the categories of environmental sounds, they were faster when the label came from transcriptions of later generation imitations than from transcriptions of first generation imitations, `r lmer_mod_results(lsn_after_first_block_lmertest_mod, "message_c")` (Fig. 7A). The effect can be further localized within each block. Comparing RTs on the trials leading up to a block transition and the trials immediately after the block transition revealed a reliable interaction between block transition and the generation of the transcribed label, `r lmer_mod_results(transition_lmertest_mod, "block_transition_c:message_c")` (Fig. 7B). In addition to becoming more stable both in terms of acoustic and orthographic properties, imitations that have been more repeated were also easier to learn as category labels.

```{r fig7, fig.height=11, out.height="11cm", fig.cap="(Top) RTs on correct trials by block, showing faster responses when learning category labels transcribed from last generation imitations. (Bottom) RTs on trials leading up to and immediately following the block transition where new category members are introduced."}
grid.arrange(rt_plot, gg_transition, ncol = 1)
```

# Discussion {-}

We've shown that repeated imitation of an originally imitative vocalization gradually becomes more word-like as it is transmitted along chains of the “Telephone” game. Imitations became more stable over generations of repetition both in terms of acoustic similarity as well as in orthographic agreement. But more than just becoming more stable over generations, the imitations also become more word-like in that they served as more effective category labels. Between-category information was more resilient to transmission decay than within-category information that identified a particular exemplar within a category. This category information remained even when the imitations were transcribed into lexical forms, as participants were able to guess the categorical meaning of the word at above chance levels even after 8 generations of repetition. One such consequence of having words as opposed to imitations is that they make categorization easier [`citations`]. In support of this conclusion, we found that participants naive to the transmission chain experiment were faster to learn category labels that had emerged through repeated imitation than they learned from transcriptions of direct imitations of the environmental sound. This evidence completes the transition from nonverbal imitation to a fully lexicalized word form and demonstrates the impact of this transition on communication.

One result that did not fit squarely with imitations becoming more word-like is that with transcriptions, there was no difference over generations between question types. If the results of matching transcriptions back to seed sounds would have perfectly mirrored the results of matching imitations back to seed sounds we would have expected the difference between True seed questions and Category match questions to decrease over generations. Instead we found a main effect of question type. Although participants were able to match transcriptions to categories of sounds even after 8 generations of repetition, it was still easier for them to match a transcription to the actual seed that generated the transcription.

Our study focused on the formation of onomatopoeia--sound-imitative words--but in addition to onomatopoeia, many languages have semantically rich systems of ideophones. These words comprise a grammatically and phonologically distinct class of words that are used to express a variety of sensory-rich meanings [@Dingemanse:2012fc; @Voeltz:2001vv]. Notably, these words are often recognized by native speakers to be somehow imitative of their meaning. For example, in Japanese, the word ‘koron’ – with a voiceless [k] refers to a light object rolling once, the reduplicated ‘korokoro’ to a light object rolling repeatedly, and ‘gorogoro’ – with a voiced [g] – to a heavy object rolling repeatedly [@Imai:2014dea]. The iconicity of ideophones was verified by an experiment that tested the ability of naïve listeners to guess the meanings of words sampled from five different languages [@Dingemanse:2016vd]. Although words for sounds were guessed more accurately than the rest, listeners were better than chance at guessing the synonyms of ideophones that expressed meanings from all five semantic categories tested -- color/visual, motion, shape, sound, and texture.  In addition, laboratory experiments show that people are able to generate imitative vocalizations for a variety of non-sound concepts, and that these are also understandable to naïve listeners [@Perlman:2015ip]. Thus vocal imitation has the potential to play a role in word formation that extends beyond just the imitation of sounds.

Our findings from an online game of Telephone suggest that the formation of words from vocal imitation can be a simple process. The results show how repeated imitation can create progressively more word-like forms while retaining a resemblance to the original sound that motivated it. This raises the possibility that onomatopoeic words can be created from the repetition of one-shot vocal imitations of an original sound.

# References {-}
