---
output_format:
  pdf_document:
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
read_chunk("creating-words.R")
```

```{r setup, include=FALSE}
```

## Selecting seed sounds

We selected inanimate categories of sounds because they were less likely to have lexicalized onomatopoeic forms already in English, and they were assumed to be less familiar and more difficult to imitate. We used an odd-one-out norming procedure (_N_ = `r n_norming_subjects` participants) to reduce an initial set of 36 sounds in 6 categories to a final set of 16 "seed" sounds: 4 sounds in each of 4 categories. The four final categories included: water, glass, tear, zipper. The results of the norming procedure are presented in the Supporting Information.

## Collecting imitations

Participants (_N_ = `r n_imitators`) were paid to participate in an online version of the children's game of "Telephone". The instructions informed participants that they would hear some sound and their task is to reproduce it as accurately as possible using their computer microphone. Full instructions are provided in the Supporting Information. Participants listened to and imitated 4 sounds. Participants received one sound from each of the four categories of sounds drawn at random such that participants were unlikely to hear the same person more than once. Recordings that were too quiet (less than `-30` dBFS) were not allowed. Imitations were monitored by an experimenter to catch any gross errors in recording before they were heard by the next generation of imitators. The experimenter also blocked sounds that violated the rules of the experiment, e.g., by saying something in English. A total of `r n_removed` imitations were removed.

## Measuring acoustic similarity

Acoustic similarity was measured by having trained research assistants listen to pairs of sounds and rate their subjective similarity. On each trial, raters heard two sounds played in succession. Then they rated the similarity between the sounds on a 7-point scale. They were instructed that a 7 on this scale meant the sounds were nearly identical, whereas a 1 meant the sounds were entirely different and would never be confused. Raters were encouraged to use as much of the scale as they could while maximizing the likelihood that, if they did this procedure again, they would reach the same judgments.

## Matching imitations to seeds

Participants (_N_ = `r n_all_matching_imitations`) listened to imitations and guessed which seeds they came from. Specifically they were instructed to select which of four possible sounds they thought the person was trying to imitate. No feedback was provided. `r n_imitation_matches_failed_catch_trial` participants failed a catch trial and were removed from the final analyses, leaving `r n_matching_imitations` participants in the final sample.

Different participants were assigned different seed sounds corresponding to the three question types depicted in Fig. 3. Participants in the True seed and Category match conditions were provided four seeds from different categories. Participants in the Specific match condition were provided four seeds from the same category. All `r n_final_imitations` were tested in each of the three conditions.

## Collecting transcriptions of imitations

Participants (_N_ = `r n_transcribers`) listened to imitations and were instructed to write down what they heard as a single word so that the written word would sound as much like the message as possible. Imitations were drawn at random from the first and last three generations of all imitations collected in the Telephone game. As a control, we also had participants "transcribe" words directly from listening to the environmental seed sounds. Transcriptions from participants who failed a catch trial were excluded (_N_ = `r n_bad_transcribers`), leaving `r n_transcriptions` for analysis. Of these, `r n_english_transcriptions` were removed because they contained English words, which was a violation of the instructions of the experiment.

## Matching transcriptions to seeds

Participants (_N_ = `r n_all_transcription_match_subjs`) completed a modified version of the "Guess the seed" game. Instead of listening to imitations, participants now read a transcription of an imitation, which they were told was an invented word. They were instructed that the word was invented to describe one of four presented sounds, and they had to guess which one. Of all the unique transcriptions that were collected for each sound (imitations and seed sounds), only the top 4 most frequent transcriptions were used in this experiment. `r n_transcription_match_subjs_failed_catch_trial` participants failed a catch trial and were excluded, leaving `r n_transcription_match_subjs` participants in the final sample.

## Learning transcriptions as category labels

Our transmission chain design and subsequent transcription procedure created `r n_created_words` novel words. From these, we sampled words transcribed from first and last generation imitations as well as from seed sounds that were equated in length and overall matching accuracy. Specifically, we removed transcriptions that contained less than 3 unique characters and transcriptions that were over 10 characters long. Of the remaining transcriptions, a sample of `r n_lsn_words` were selected to have approximately equal means and variances of overall matching accuracy.

Participants (_N_ = `r n_all_lsn_subjs`) were randomly assigned four novel names for four categories of environmental sounds. They learned the referents for these names in a trial-and-error category learning experiment. On each trial, participants heard one of the 16 seed sounds and then saw a word--one of the transcriptions of the imitations. They responded yes or no using a gamepad as to whether the sound and the word went together. Initially they were forced to guess, but because they received feedback on their performance, over trials they learned the names of the categories. `r n_lsn_subjs` outlier participants were excluded from the final sample due to high error rates and slow reaction times.

Participants categorized all 16 seed sounds over the course of the experiment, but they learned them in blocks of 4 sounds at a time. Within each block, participants heard the same four sounds and the same four words multiple times, with a 50% probability of the sound matching the word. At the start of a new block of trials, participants heard four new sounds they hadn't heard before, and had to learn to associate these new sounds with the words they had learned in the previous blocks.
