---
output_format:
  pdf_document:
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
read_chunk("creating-words.R")
```

```{r setup, include=FALSE}
```

## Selecting seed sounds

We selected inanimate categories of sounds because they were less likely to have lexicalized onomatopoeic forms already in English, and they were assumed to be less familiar and more difficult to imitate. We used an odd-one-out norming procedure (_N_ = `r n_norming_subjects` participants) to reduce an initial set of 36 sounds in 6 categories to a final set of 16 "seed" sounds: 4 sounds in each of 4 categories. The four final categories included: water, glass, tear, zipper. The results of the norming procedure are presented in the Supporting Information.

## Collecting imitations

Participants (_N_ = `r n_imitators`) were paid to participate in an online version of the children's game of "Telephone". The instructions informed participants that they would hear some sound and their task is to reproduce it as accurately as possible using their computer microphone. Full instructions are provided in the Supporting Information. Participants listened to and imitated 4 sounds. Participants received one sound from each of the four categories of sounds drawn at random such that participants were unlikely to hear the same person more than once. Recordings that were too quiet (less than `-30` dBFS) were not allowed. Imitations were monitored by an experimenter to catch any gross errors in recording before they were heard by the next generation of imitators. The experimenter also blocked sounds that violated the rules of the experiment, e.g., by saying something in English. A total of `r n_removed` imitations were removed.

## Matching imitations to seeds

Participants (_N_ = `r n_matching_imitations`) listened to imitations and guessed which seeds they came from. 

## Collecting transcriptions of imitations

Participants were instructed to write down what they heard as a word so that the written word would sound as much like the message as possible.

## Matching transcriptions to seeds

Instead of listening to imitations, participants now read a transcription of an imitation, which they were told was an invented word. They were instructed that the word was invented to describe one of four presented sounds, and they had to guess which one.

## Learning transcriptions as category labels
