---
title: The emergence of words from iterated vocal imitation
author: |
  Pierce Edmiston, Marcus Perlman, and Gary Lupyan  
  University of Wisconsin-Madison
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results="hide", cache=TRUE, cache.path="talk_cache/")
read_chunk("analysis.R", labels = "main")
```

```{r main, include=FALSE}
```

# Are names for things natural or conventional?

**Hermogenes**: Conventional!  
Names like "man" and "horse" are interchangeable.  
**Cratylus**: Natural!  
_anthrôpos_ = "man" ~ _anathrôn ha opôpe_ = "one who reflects on what he has seen".  (Some horse!)

```{r plato}
draw_image("img/plato.jpg", height = 0.2)
```

<!--
Are names for things natural or conventional?

Are some names more true or correct than others,
or are all names to some extent interchangeable?

This was the question that Plato explored in his _Cratylus_ dialogue.
The dialogue is a debate between two characters, Cratylus and Hermogenes.

Hermogenes argues that names are conventional. We could just as
easily call "man" "horse", and "horse" "man".

Cratylus argues that names are natural;
they can be either correct or incorrect.

As an example, the Greek word for man can be traced to its
roots to mean "one who reflects on what he has seen", and
so to Cratylus it seems ridiculous to think that the word
for man could just as easily mean "horse".
-->

# What about the first names?

> **Socrates**: If we wished to designate that which is above and is light, we should raise our hands towards heaven in imitation; but if the things to be designated were below or heavy, we should extend our hands towards the ground...

> A name, then, it appears, is a vocal imitation of that which is imitated, and he who imitates with his voice names that which he imitates.

> **Hermogenes**: But Socrates, what sort of an imitation is a name?

<!--
Putting aside etymologies,
where new names are derived from old names,
we can also consider the correctness of names without any precedents:
the first names.

So what were the first names, were they natural or conventional?

Here, Socrates, who's mediating this debate,
sides with Cratylus. First he points out that imitation 
is a natural way to designate something without prior convention.

[read first quote]

He goes on to say that the same is true not just for bodily imitations,
but for vocal imitations as well.

[read second quote]

But then Hermogenes pushes back a little.

[read third quote]

How do you get from an imitation _of_ something to a name _for_ something?

Then they go back and forth for awhile, ultimately conceding that some
names seem more natural than others, but for the most part, the debate
isn't fully settled, which is why I'm still talking about it a couple
thousand years laters.
-->

# Which is more important to language evolution?

```{r lupyanlab-dialogue, engine="dot"}
digraph {
  rankdir=LR;
  node[shape="none" fontname=Helvetica fontcolor=white fontsize=20];
  edge[arrowhead="vee" arrowtail="vee"];
  marcus -> pierce[dir=both minlen=1.5 style=invis];
  marcus[label="Iconicity!" image="img/marcus.png" labelloc=b];
  pierce[label="Arbitrariness!" image="img/pierce.png" labelloc=b];
}
```

**Arbitrariness**: facilitates categorization, but difficult to learn.  
**Iconicity**: easy to learn, but doesn't facilitate categorization.

<!--
I bring this up because it's reminiscent of a similar dialogue I
had with Marcus Perlman, back when we started this project.

The debate we had was this:

Which is a more important contributor to language evolution?
Iconicity,
or what Plato would have called the "naturalness of some names",
or arbitrariness,
the idea that words do not have to resemble their meanings?

In the past, I've argued for the arbitrariness side of this equation.
What I've said is that
one of the reason language is so advantageous
is not because it allows us to refer to specific things,
but that it allows us to group together categories of things,
irrespective of the individual members.

Through arbitrariness, you get categorization and symbolic reasoning.

In contrast, Marcus has argued for the role of iconicity and vocal imitation
in language evolution, and he's pointed out that language isn't
perfectly arbitrary, also that people are very good at both generating
and interpreting vocal imitations without prior convention.

So Marcus, Gary, and I got together to try to figure out these differences.
Specifically what we were after is a way to see between
this dichotomy, to see the advantages of iconicity when creating
words, but also to observe the advantages of more symbolic
mappings for things like categorization.
-->

# Does repeating imitations make them more wordlike?

```{r telephone-chain, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname=Helvetica shape=circle];
  0 -> 1 -> 2 -> 3 -> 4 -> 5;
  0[shape="none" label="imitation"];
  5[shape="none" label="word?"];
}
```

<!--
Specifically what we asked was whether words can emerge from
iterated imitations.

We were interested in the minimal conditions for observing
an emergence of words from imitations.
Without any instruction or intention to create a word,
does simply repeating it cause it to become more wordlike?

To test this we had participants imitate various environmental sounds,
sounds like glass breaking, and water splashing,
and then we had new participants imitate the imitations,
and so on for generation after generation.

What we measured was how the imitation changed as a function of
generation of repetition.

Is repeating an imitation over and over across generations of different
speakers sufficient to make it more wordlike?
-->

# Measuring wordlike-ness

1. Stable
2. Learned
3. Categorical

# Interface for collecting vocal imitations

```{r telephone-game}
crotchet::draw_image("collect-imitations-gui", "wordsintransition")
```

<!--
Here's what the experiment looked like.

It was basically our version of the children's game of telephone.

Participants clicked on the top speaker to hear a sound,
and the bottom speaker to make their imitation of it.

And participants were instructed to copy whatever they heard as accurately as possible.
-->

# The results of the transmission chain experiment

```{r}
gg_dendrogram
```

<!--
And here are our results -- the imitations we collected in this
transmission chain experiment.

We had four different categories of environmental sounds:
water, glass, tearing, and zipper,
and four seed sounds within each category.

The first generation of participants imitated the seed sound,
and later generations imitated the previous generation,
for a maximum of 8 generations.
-->

# Research questions

1. Do repeated imitations stabilize on particular words?
2. Do the imitations retain their resemblance to the seed sound?
3. Do the imitations become more suitable as category labels?

<!--
Once we collected these imitations, we sought to answer three questions
about how the imitations changed as a function of repetition.

First, do the imitations in each chain stabilize on particular words?

Are later generation imitations more similar to one another than earlier
generation imitations?

Next we asked what the imitations were stabilizing on. Do the imitations
retain a resemblance to the original seed sound, or does each chain
careen off to create a total nonsence word.

Finally, we asked whether repetition has any reprecussions for the
language learner. We took words created at the beginning of the
chains and at the end of the chains, and test them to see which served
as better category labels in a category learning experiment.
-->

# Repeated imitations became more similar

```{r similarity-judgments}
gg_similarity_judgments
```

# Similarity increased within, not between, chains

```{r algorithmic-similarity}
gg_algo_similarity
```

# Example transcriptions

```{r example-transcriptions, results="asis"}
kable(transcription_examples)
```

# Repeated imitations were transcribed more consistently

```{r transcription-distance}
gg_distance
```

# 1. Do imitations stabilize on particular words?

**Yes.** Repeating imitations makes them more similar to one another acoustically and orthographically.

_But what are they stabilizing on?_

# Interface for matching imitations to seed sounds

```{r guess-the-seed}
draw_image("match-imitations-gui", "wordsintransition")
```

# Types of matching questions

```{r matching-questions}
grob_question_types <- arrangeGrob(
  read_graphviz("true-seed", "wordsintransition"),
  read_graphviz("category-match", "wordsintransition"),
  read_graphviz("specific-match", "wordsintransition"),
  ncol = 1
)
grid::grid.draw(grob_question_types)
```

# Category information was the most resilient to decay

```{r matching-imitations}
grid.arrange(
  grob_question_types,
  gg_match_to_seed,
  nrow = 1
)
```

# Transcriptions can retain category informations

```{r matching-transcriptions}
gg_match_transcriptions
```

# 2. Do imitations retain a resemblance to the seed sound?

**Yes** (at least for 8 generations with 4 categories).

_What are the consequences of repeating imitations for learners?_

# A simple category learning experiment

```{r learning-sound-names-paradigm, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname="Helvetica" shape=none];
  sound -> label -> prompt[minlen=2];
  sound[label="sound\n<water>\n 1/16"];
  label[label="word\n'eeverlusha'\n 1/4"];
  prompt[label="?\nY/N"];
}
```

# No difference in accuracy for novel labels

```{r learning-performance-ceiling}
gg_lsn_performance_ceiling
```

# Later generation labels yielded faster responses

```{r learning-rts}
rt_plot
```

# Later generation labels were generalized more quickly

```{r learning-generalization}
gg_transition
```

# The emergence of words from iterated vocal imitation

1. Do repeated imitations stabilize on particular words?
2. Do the imitations retain their resemblance to the seed sound?
3. Do the imitations become more suitable as category labels?

# Thanks!

Pierce Edmiston, Marcus Perlman, and Gary Lupyan.
[github.com/lupyanlab/creating-words](https://github.com/lupyanlab/creating-words/tree/cesc)  
[osf.io/3navm](https://osf.io/3navm)  

# References

- http://www.hermes-press.com/cratylus2.htm
- https://plato.stanford.edu/entries/plato-cratylus/

# Correlation between subjective and objective similarity

```{r comparing-similarities}
gg_comparing_similarities
```

# Transcriptions of seeds were the least consistent

```{r transcription-of-seeds-distance}
gg_seed_distance
```

# Match accuracies for imitations and transcriptions

```{r matching-accuracies}
grid.arrange(
  gg_match_to_seed,
  gg_match_transcriptions + theme(axis.title.y = element_blank()),
  nrow = 1
)
```

# Transcriptions of seed sounds were matched very accurately

```{r matching-transcriptions-of-seeds}
gg_seed_matching
```

# Transcriptions of seed sounds ???

```{r learning-transcriptions-of-seeds}
gg_seed_rt_plot
```

# Onomatopoeic words are found across languages

```{r onomatopoeia-comics, fig.cap="By James Chapman."}
gridExtra::grid.arrange(
  crotchet::read_image("img/chapman-clap.png"),
  crotchet::read_image("img/chapman-sneeze.png"),
  nrow = 1
)
```

