---
title: The emergence of words from iterated vocal imitation
author: Pierce Edmiston, Marcus Perlman, and Gary Lupyan
output:
  beamer_presentation:
    theme: metropolis
---

# How good are humans at vocal imitation?

```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results="hide", cache=TRUE)
sapply(list.files("R", "*.R", full.names = TRUE), read_chunk)
```

```{r setup}
```

```{r stability}
```

```{r matching}
```

```{r learning}
```

# The design of the transmission chain experiment

```{r}
gg_dendrogram
```

# Repeating imitations increases acoustic similarity

```{r algorithmic-similarity}
gg_algo_similarity +
  theme(legend.position = c(0.45, 0.9))
```

# Similarity increases within, not between, chains

```{r algorithmic-similarity-types}
gg_algo_compare
```

# Repeated imitations are transcribed more consistently

```{r transcription-distance}
gg_distance
```

# Types of matching questions

```{r matching-questions}
grid.arrange(
  read_graphviz("true-seed", "wordsintransition"),
  read_graphviz("category-match", "wordsintransition"),
  read_graphviz("specific-match", "wordsintransition"),
  ncol = 1
)
```

# Category information was the most resilient to decay

```{r matching-imitations}
gg_match_to_seed
```

# Transcriptions can retain category information

```{r matching-transcriptions}
gg_match_transcriptions
```

# Later generation labels yielded faster responses

```{r learning-rts}
rt_plot +
  theme(legend.position = c(0.55, 0.9))
```

# No difference in accuracy for novel labels

```{r learning-performance-ceiling}
gg_lsn_performance_ceiling +
  theme(legend.position = c(0.45, 0.9))
```

# Later generation labels were generalized more quickly

```{r learning-generalization}
gg_transition +
  theme(legend.position = c(0.45, 0.9))
```

# The emergence of words from iterated vocal imitation

<!----------------------------------------------------------------------------->

# Correlation between subjective and objective similarity

```{r comparing-similarities}
gg_comparing_similarities
```

# Transcriptions of seeds were the least consistent

```{r transcription-of-seeds-distance}
gg_seed_distance
```

# Match accuracies for imitations and transcriptions

```{r matching-accuracies}
grid.arrange(
  gg_match_to_seed,
  gg_match_transcriptions + theme(axis.title.y = element_blank()),
  nrow = 1
)
```

# Transcriptions of seed sounds were matched very accurately

```{r matching-transcriptions-of-seeds}
gg_seed_matching
```

# Transcriptions of seed sounds ???

```{r learning-transcriptions-of-seeds}
gg_seed_rt_plot +
  theme(legend.position = c(0.55, 0.9))
```
