---
title: The emergence of words from iterated vocal imitation
author: |
  Pierce Edmiston, Marcus Perlman, and Gary Lupyan  
  University of Wisconsin-Madison
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results="hide", cache=TRUE)
read_chunk("analysis.R", labels = "main")
```

```{r main, include=FALSE}
```

# Are names for things natural or conventional?

```{r plato}
draw_image("img/plato.jpg", height = 0.3)
```

<!--
I want to start with a very old question about language, and
about where names come from, and that is,
are names for things natural or are they conventional?

Are some names better or more natural than others, or
are all names created equal and interchangeable?

This is the question that Plato explored in his _Cratylus_
dialogue, and his answer was basically that both positions
are wrong in the extremes.

While some names seem better or more natural than others,
there could not be just one ideal name for every thing,
and so convention must play a role in naming.
-->

# Shakespeare on names

<!--
This debate has carried on for many hundreds of years.
Here's Shakespeare commenting on the naturalness of
names in _Romeo and Juliet_, where he seems to side
with the conventional side of the debate, that
"a rose by any other name would still sound as sweet".
-->

> What's in a name? that which we call a rose  
> By any other word would smell as sweet;

# Which aspect of language is more important?

```{r marcus}
draw_image("img/marcus.png", height = 0.3)
```

# Onomatopoeic words are found across languages

```{r onomatopoeia-comics, fig.cap="By James Chapman."}
gridExtra::grid.arrange(
  crotchet::read_image("img/chapman-clap.png"),
  crotchet::read_image("img/chapman-sneeze.png"),
  nrow = 1
)
```

```{r telephone-game}
crotchet::draw_image("collect-imitations-gui", "wordsintransition")
```

```{r telephone-chain, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname=Helvetica shape=circle];
  1 -> 2 -> 3 -> 4 -> 5;
  5[style=invis];
}
```

# The design of the transmission chain experiment

```{r}
gg_dendrogram
```

<!----------------------------------------------------------------------------->

# Repeating imitations increased acoustic similarity

```{r similarity-judgments}
gg_similarity_judgments
```

# Similarity increased within, not between, chains

```{r algorithmic-similarity}
gg_algo_similarity
```

# Repeated imitations were transcribed more consistently

```{r transcription-distance}
gg_distance
```

<!----------------------------------------------------------------------------->

# Recap

# Types of matching questions

```{r matching-questions}
grid.arrange(
  read_graphviz("true-seed", "wordsintransition"),
  read_graphviz("category-match", "wordsintransition"),
  read_graphviz("specific-match", "wordsintransition"),
  ncol = 1
)
```

# Category information was the most resilient to decay

```{r matching-imitations}
gg_match_to_seed
```

# Transcriptions can retain category information

```{r matching-transcriptions}
gg_match_transcriptions
```

<!----------------------------------------------------------------------------->

# Recap

# Later generation labels yielded faster responses

```{r learning-rts}
rt_plot
```

# No difference in accuracy for novel labels

```{r learning-performance-ceiling}
gg_lsn_performance_ceiling
```

# Later generation labels were generalized more quickly

```{r learning-generalization}
gg_transition
```

# The emergence of words from iterated vocal imitation

<!----------------------------------------------------------------------------->

# Correlation between subjective and objective similarity

```{r comparing-similarities}
gg_comparing_similarities
```

# Transcriptions of seeds were the least consistent

```{r transcription-of-seeds-distance}
gg_seed_distance
```

# Match accuracies for imitations and transcriptions

```{r matching-accuracies}
grid.arrange(
  gg_match_to_seed,
  gg_match_transcriptions + theme(axis.title.y = element_blank()),
  nrow = 1
)
```

# Transcriptions of seed sounds were matched very accurately

```{r matching-transcriptions-of-seeds}
gg_seed_matching
```

# Transcriptions of seed sounds ???

```{r learning-transcriptions-of-seeds}
gg_seed_rt_plot
```
