---
title: The emergence of words from iterated vocal imitation
author: |
  Pierce Edmiston, Marcus Perlman, and Gary Lupyan  
  University of Wisconsin-Madison  
  pedmiston@wisc.edu
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results="hide", cache=TRUE, cache.path="talk_cache/")
read_chunk("analysis.R", labels = "main")
```

```{r main, include=FALSE}
```

# Are names for things natural or conventional?

Plato's _Cratylus_

```{r plato}
draw_image("img/plato.jpg", height = 0.2)
```

**Hermogenes**: Conventional! Names like "man" and "horse" are arbitrary.  
**Cratylus**: Natural! _anthrôpos_ = "man" ~ _anathrôn ha opôpe_ = "one who reflects on what he has seen". (Some horse!)

<!--
Are names for everyday things natural or conventional?

Are some names better or more correct than others,
or are all names to some extent interchangeable?

This was the question that Plato explored in his _Cratylus_ dialogue,
which features a debate between Cratylus and Hermogenes.

Hermogenes argues that names are conventional and arbitrary.
We could just as easily call "man" "horse" as "horse" "man".

Cratylus argues the opposite, that names are natural;
some names are correct, and others are incorrect.
"Man" is the correct name for "man",
and "horse" is the correct name for horses.

Socrates, who's mediating this debate,
backs up Cratylus by giving examples like this:
he traces the Greek word for "man"
to mean "one who reflects on what he has seen",
and concludes, well, if we called a horse that,
it would have to be one hell of a horse!
-->

# What were the first words?

> **Socrates**: If we wished to designate that which is above and is light, we should raise our hands towards heaven in imitation; but if the things to be designated were below or heavy, we should extend our hands towards the ground...

> A name, then, it appears, is a vocal imitation of that which is imitated, and **he who imitates with his voice names that which he imitates.**

> **Hermogenes**: But Socrates, **what sort of an imitation is a name?**

<!--
Creating new words out of existing words is one way names can be natural,
but Plato also explored whether words can be formed naturally,
without any prior convention,
and this is the part that I'm going to focus on today.

The question is whether the first words had to be conventional,
or whether they might have been natural as well.

Here, Socrates is siding with Cratylus.
First he points out that
imitation is a natural way to designate something
without prior convention.

> If we wished to designate that which is above and is light, we should raise our hands towards heaven in imitation; but if the things to be designated were below or heavy, we should extend our hands towards the ground.

He goes on to say that this applies to vocal imitations as well.

> A name, then, it appears, is a vocal imitation of that which is imitated, and **he who imitates with his voice names that which he imitates.**

But then Hermogenes pushes back a little,

> But Socrates, what sort of an imitation is a name?

He accepts that imitations can be used as names,
but given a name, he doesn't see the imitation.
-->

# Central question

How do you get from an imitation _of_ something to a name _for_ something?

<!--
That brings us to the central question of this research:

How do you get from an imitation _of_ something
to a name _for_ something?

Does it require a conscious decision to invent or assign a new name?
Or might imitations lead naturally to new names?
-->

# Vocal imitations are not words

- Imitations are idiosyncratic, words are **stable**.
- Imitations are specific, words are **categorical**.

<!--
By asking this question,
I want to make clear the differences between vocal imitations
and conventional words.

There are two dimensions along which imitations are different from words.

The first is that imitations are idiosyncratic.
If I asked everyone here to imitate a dog,
it's likely that each person's imitation of a bark
will be more different from all the others
than each person's pronunciation of the word "dog",
which is much more stable across individuals.

The second is that imitations are specific.
An imitation of a dog bark is an imitation of a specific dog,
at a specific time, and in a specific emotional state;
whether the dog is being aggressive or submissive.
Imitations are recreations of specific environmental events.

Words, on the other hand, leave all of that unspecified.
The word "dog" doesn't tell you anything about how big the dog is,
or whether the dog is playing or fighting,
only that it is a dog.

Words are categorical.
They refer to all members of the category equally.
-->

# How do you turn an imitation into a word?

```{r telephone-chain, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname=Helvetica shape=circle];
  0 -> 1 -> 2 -> 3 -> 4 -> 5 -> 6 -> 7 -> 8;
  0[shape="none" label="sound"];
  8[shape="none" label="word?"];
}
```

<!--
Now that we know the two dimensions along which imitations differ from words,
we can ask how you might get from an imitation to a word.

Our idea,
which is not an original idea by any means,
but we seem to be the first to try to test it formally,
was that simply repeating an imitation
across unique individuals
might be sufficient
to make imitations
more stable,
more categorical,
and therefore more wordlike,
even without explicit intention to conventionalize.
-->

# Interface for collecting vocal imitations

```{r telephone-game}
crotchet::draw_image("collect-imitations-gui", "wordsintransition")
```

<!--
To test this,
we ran an online version of the children's game of Telephone,
where a message is passed from person to person with the goal of staying unchanged.

Participants logged in to the experiment,
and were instructed they would be playing something like Telephone,
but instead of a normal sentence or phrase,
they would hear some sound,
like glass breaking or water splashing,
or someone imitating of one of these sounds.

We told them that no matter how weird the sound was,
their task was exactly the same as in the children's game:
to repeat whatever they heard as accurately as possible.

Let me play you some examples just so you get a sense of what this sounded like.

<seed, gen1>
<seed, gen1>
<seed, gen1-gen8>
<seed, gen1-gen8>
<seed, gen1-gen8>
-->

# The results of the transmission chain experiment

```{r}
gg_dendrogram
```

<!--
Overall we collected about 500 imitations
but from these,
we had to remove about a quarter,
because they were too quiet,
or there was too much background noise,
or they violated our instructions by saying something in English,
as in the "a lobster" case.

The final imitations are shown here:
365 imitations along 105 contiguous transmission chains.
-->

# Research questions

Do repeated imitations:

1. Stabilize on particular words?
2. Retain their resemblance to the seed sound?
3. Become more suitable as category labels?

<!--
Once we collected these imitations, we sought to answer three questions
about how the imitations changed as a function of being repeated.

First, do the imitations in each chain stabilize on particular words?
Are later generation imitations
more similar to one another
than earlier generation imitations?

Next we asked what the imitations were stabilizing on.
Do the imitations retain a resemblance to the original seed sound,
or does each chain careen off to create a total nonsense word.

Finally, we asked whether repetition has any consequences for the language learner.
We took words from the beginning and the end of the chains,
and tested whether repetition made the words easier to learn as category labels and easier to generalize to new category members.
-->

# Repeated imitations became more similar

```{r similarity-judgments}
gg_similarity_judgments
```

<!--
First up is whether the imitations
became more stable as they were repeated.

To test this,
we had people rate the similarity between adjacent sounds
within each transmission chain.

We found that later generation imitations
were rated as more similar to one another
than earlier generation imitations,
supporting our idea that repetition
might make the imitations
more stable and repeatable.
-->

# Similarity increased within, not between, chains

```{r algorithmic-similarity}
gg_algo_similarity
```

<!--
An alternative possibility
is that the imitations were becoming more similar
not because they were stabilizing,
but because any acoustic signal
that's repeated from person to person
might degrade to the point
that they all sound the same.

To test this,
we compared the change in acoustic similarity within chains
to an analogous comparison between chains,
the idea being
that if continuous degradation of the signal
was what was causing the increase in similarity,
that similarity might increase between chains as well as within.

Instead we found that similarity only increased within chains,
and not between them.
-->

# Example transcriptions

```{r example-transcriptions, results="asis"}
kable(transcription_examples)
```

<!--
As a final measure of stability,
we looked at how people's transcriptions of the imitations
changed as a function of generation.

We had people listen to imitations
sampled from the first and last generation of every chain
and transcribe it into a word.

We instructed them to write down what they heard as a word
that sounded as much like the sound as possible.

Here are some examples:
-->

# Repeated imitations were transcribed more consistently

```{r transcription-distance}
gg_distance
```

<!--
Next we measured the agreement among the transcriptions
using a string distance measure
and found that later generation imitations
were transcribed more consistently
than earlier generation imitations.
-->

# 1. Do imitations stabilize on particular words?

**Yes.** Repeating imitations makes them more similar to one another acoustically and orthographically.

_But what are they stabilizing on?_

<!--
So, does repeating an imitation cause it to stabilize on particular words?

Yes. We found that repeated imitations were more similar to one another
acoustically as well as orthographically.

However, we don't know what the imitations are stabilizing on,
and whether and for how long the imitations retain a resemblance
to the original seed sound.
-->

# Interface for matching imitations to seed sounds

```{r guess-the-seed}
draw_image("match-imitations-gui", "wordsintransition")
```

<!--
To test this,
we had people listen to each of the imitations,
and four of the different seed sounds,
and they had to guess
which sound they thought the person was trying to imitate.
-->

# Types of matching questions

```{r matching-questions}
grob_question_types <- arrangeGrob(
  read_graphviz("true-seed", "wordsintransition"),
  read_graphviz("category-match", "wordsintransition"),
  read_graphviz("specific-match", "wordsintransition"),
  ncol = 1
)
grid::grid.draw(grob_question_types)
```

# Category information was the most resilient to decay

```{r matching-imitations}
grid.arrange(
  grob_question_types,
  gg_match_to_seed,
  nrow = 1
)
```

# Transcriptions can retain category informations

```{r matching-transcriptions}
gg_match_transcriptions
```

# 2. Do imitations retain a resemblance to the seed sound?

**Yes** (at least for 8 generations with 4 categories).

_What are the consequences of repeating imitations for learners?_

# Selecting words to use as category labels

- Filtered for length and number of unique characters.
- Equated based on overall matching accuracy.
- Selected **a subset** (56) for testing.

# A simple category learning experiment

- Learned to associate 4 novel labels with 4 categories of sounds.
- Categorized 16 sounds total, 4 per block of trials.
- Trial-and-error learning procedure.

```{r learning-sound-names-paradigm, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname="Helvetica" shape=none];
  sound -> label -> prompt[minlen=2];
  sound[label="sound\n<water 2>"];
  label[label="word\n'eeverlusha'"];
  prompt[label="?\nY/N"];
}
```

# Later generation labels yielded faster responses

```{r learning-rts}
rt_plot
```

# No difference in accuracy for novel labels

```{r learning-performance-ceiling}
gg_lsn_performance_ceiling
```

# Later generation labels were generalized more quickly

```{r learning-generalization}
gg_transition
```

# The emergence of words from iterated vocal imitation

1. Do repeated imitations stabilize on particular words?
2. Do the imitations retain their resemblance to the seed sound?
3. Do the imitations become more suitable as category labels?

# Thanks!

Pierce Edmiston, Marcus Perlman, and Gary Lupyan.
[github.com/lupyanlab/creating-words](https://github.com/lupyanlab/creating-words/tree/cesc)  
[osf.io/3navm](https://osf.io/3navm)  

# References

- http://www.hermes-press.com/cratylus2.htm
- https://plato.stanford.edu/entries/plato-cratylus/

# Correlation between subjective and objective similarity

```{r comparing-similarities}
gg_comparing_similarities
```

# Transcriptions of seeds were the least consistent

```{r transcription-of-seeds-distance}
gg_seed_distance
```

# Match accuracies for imitations and transcriptions

```{r matching-accuracies}
grid.arrange(
  gg_match_to_seed,
  gg_match_transcriptions + theme(axis.title.y = element_blank()),
  nrow = 1
)
```

# Transcriptions of seed sounds were matched very accurately

```{r matching-transcriptions-of-seeds}
gg_seed_matching
```

# Transcriptions of seed sounds ???

```{r learning-transcriptions-of-seeds}
gg_seed_rt_plot
```

# Onomatopoeic words are found across languages

```{r onomatopoeia-comics, fig.cap="By James Chapman."}
gridExtra::grid.arrange(
  crotchet::read_image("img/chapman-clap.png"),
  crotchet::read_image("img/chapman-sneeze.png"),
  nrow = 1
)
```
