---
title: The emergence of words from iterated vocal imitation
author: |
  Pierce Edmiston, Marcus Perlman, and Gary Lupyan  
  University of Wisconsin-Madison
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: true
---

```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results="hide", cache=TRUE)
read_chunk("analysis.R", labels = "main")
```

```{r main, include=FALSE}
```

# Are names for things natural or conventional?

```{r cratylus-dialogue, engine="dot"}
digraph {
  rankdir=LR;
  fontname=Helvetica;
  label="Plato's Cratylus";
  labelloc=t;
  node[shape="none" fontname=Helvetica];
  edge[arrowhead="vee" arrowtail="vee"];
  Cratylus -> Hermogenes[dir=both minlen=2 shape=circle];
  Cratylus[label="Cratylus\n'Natural!'"];
  Hermogenes[label="Hermogenes\n'Conventional!'"];
}
```

<!--
I want to start with a very old question about language, and
about where names come from, and that is,
are names for things natural or are they conventional?

Are some names better than others, or
are all names created equal and to some extent interchangeable?

This is a question that Plato explored in his _Cratylus_ dialogue. The
dialogue is an argument between Cratylus and Hermogenes.

Cratylus argued that names are not arbitrarily decided because some names seem
better or more natural than others.

Hermogenes argued the exact opposite,
that only convention could determine which names are used to designate
which objects.

The dialogue goes on to argue that both views are wrong in the extremes,
but lucky for us, they didn't solve it, and this question has been
the topic of much discussion for many hundreds of years.
-->

# Which is more important to language evolution?

```{r lupyanlab-dialogue, engine="dot"}
digraph {
  rankdir=LR;
  node[shape="none" fontname=Helvetica fontcolor=white fontsize=20];
  edge[arrowhead="vee" arrowtail="vee"];
  marcus -> pierce[dir=both minlen=2 shape=none];
  marcus[label="Iconicity!" image="img/marcus.png" labelloc=b];
  pierce[label="Arbitrariness!" image="img/pierce.png" labelloc=b];
}
```

- Arbitrary names promote categorization (Edmiston & Lupyan, 2015).
- Iconic names are easier to learn (Perry, Perlman, & Lupyan, 2016).

<!--
I bring this up because it's reminiscent of a similar dialogue I
had with Marcus Perlman, back when we started this project.

The debate we had was this:

Which is a more important contributor to language evolution?
Iconicity,
or what Plato might have called the "naturalness of some names",
or arbitrariness,
the idea that words do not have to resemble their meanings?

In the past, I've argued for the arbitrariness side of this equation.
What I've said is that
the reason language is so advantageous
is not because it allows us to refer to specific things,
but that it allows us to form categories,
irrespective of the individual members.

Through arbitrariness, you get categorization and symbolic reasoning.

Marcus believes something a little different. He thinks that the
widespread belief that languages today are entirely arbitrary was bogus.
In fact, iconicity pervades languages, even ones like English and
Spanish that are thought to be arbitrary.

The reason widespreaad iconicity would be important for language evolution
is that iconic names may be easier to learn.
In a project with Lynn Perry,
they found was that words that were rated
as more iconic were also learned earlier in a child's life.

So we needed a way to settle our debate, and we decided to do
an experiment.
-->

# Onomatopoeic words are found across languages

```{r onomatopoeia-comics, fig.cap="By James Chapman."}
gridExtra::grid.arrange(
  crotchet::read_image("img/chapman-clap.png"),
  crotchet::read_image("img/chapman-sneeze.png"),
  nrow = 1
)
```

```{r telephone-game}
crotchet::draw_image("collect-imitations-gui", "wordsintransition")
```

```{r telephone-chain, engine="dot"}
digraph {
  rankdir=LR;
  node[fontname=Helvetica shape=circle];
  1 -> 2 -> 3 -> 4 -> 5;
  5[style=invis];
}
```

# The transmission chain experiment

```{r}
gg_dendrogram
```

<!----------------------------------------------------------------------------->

# Repeating imitations increased acoustic similarity

```{r similarity-judgments}
gg_similarity_judgments
```

# Similarity increased within, not between, chains

```{r algorithmic-similarity}
gg_algo_similarity
```

# Repeated imitations were transcribed more consistently

```{r transcription-distance}
gg_distance
```

<!----------------------------------------------------------------------------->

# Recap

# Types of matching questions

```{r matching-questions}
grid.arrange(
  read_graphviz("true-seed", "wordsintransition"),
  read_graphviz("category-match", "wordsintransition"),
  read_graphviz("specific-match", "wordsintransition"),
  ncol = 1
)
```

# Category information was the most resilient to decay

```{r matching-imitations}
gg_match_to_seed
```

# Transcriptions can retain category information

```{r matching-transcriptions}
gg_match_transcriptions
```

<!----------------------------------------------------------------------------->

# Recap

# Later generation labels yielded faster responses

```{r learning-rts}
rt_plot
```

# No difference in accuracy for novel labels

```{r learning-performance-ceiling}
gg_lsn_performance_ceiling
```

# Later generation labels were generalized more quickly

```{r learning-generalization}
gg_transition
```

# The emergence of words from iterated vocal imitation

<!----------------------------------------------------------------------------->

# Correlation between subjective and objective similarity

```{r comparing-similarities}
gg_comparing_similarities
```

# Transcriptions of seeds were the least consistent

```{r transcription-of-seeds-distance}
gg_seed_distance
```

# Match accuracies for imitations and transcriptions

```{r matching-accuracies}
grid.arrange(
  gg_match_to_seed,
  gg_match_transcriptions + theme(axis.title.y = element_blank()),
  nrow = 1
)
```

# Transcriptions of seed sounds were matched very accurately

```{r matching-transcriptions-of-seeds}
gg_seed_matching
```

# Transcriptions of seed sounds ???

```{r learning-transcriptions-of-seeds}
gg_seed_rt_plot
```
